---
title: "Measurement of sex and gender identity in BRFSS"
author: "Jenine K Harris, Odalis Hernandez"
date: "9/15/2021"
output: word_document
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
# TO CREATE ANALYTIC DATA SET, UNCOMMENT AND RUN THIS CODE CHUNK
# bring in files from online BRFSS 2015 to 2018 data
library(haven)
# #
# #
# # # 2015 data
  temp <- tempfile(fileext = ".zip")
  download.file(url  = "https://www.cdc.gov/brfss/annual_data/2015/files/LLCP2015XPT.zip", destfile = temp)
  BRFSS_2015 <- read_xpt(file = temp)
# #
# # 2016 data
 temp <- tempfile(fileext = ".zip")
 download.file(url  = "https://www.cdc.gov/brfss/annual_data/2016/files/LLCP2016XPT.zip", destfile = temp)
 BRFSS_2016 <- read_xpt(file = temp)
#
# # 2017 data
 temp <- tempfile(fileext = ".zip")
 download.file(url  = "https://www.cdc.gov/brfss/annual_data/2017/files/LLCP2017XPT.zip", destfile = temp)
 BRFSS_2017 <- read_xpt(file = temp)
#
# 2018 data
temp <- tempfile(fileext = ".zip")
download.file(url  = "https://www.cdc.gov/brfss/annual_data/2018/files/LLCP2018XPT.zip", destfile = temp)
BRFSS_2018 <- read_xpt(file = temp)

# 2019 data
temp <- tempfile(fileext = ".zip")
download.file(url  = "https://www.cdc.gov/brfss/annual_data/2019/files/LLCP2019XPT.zip", destfile = temp)
BRFSS_2019 <- read_xpt(file = temp)

# 2020 data
temp <- tempfile(fileext = ".zip")
download.file(url  = "https://www.cdc.gov/brfss/annual_data/2020/files/LLCP2020XPT.zip", destfile = temp)
BRFSS_2020 <- read_xpt(file = temp)
```

```{r}
#
# # select only the variables to use in the analysis (e.g. changes 300+ variables to only needed for analysis)

#
library(tidyverse)
 BRFSS_2015 <- BRFSS_2015 %>%
   select(SEX, TRNSGNDR, `_RFBING5`, `_RFDRHV5`, DIABETE3,
          PREGNANT, HADMAM, HOWLONG, HADPAP2, LASTPAP2, HPVTEST,
          HPLSTTST, HADHYST2, PROFEXAM, LENGEXAM, PCPSAAD2, PCPSADI1,
          PCPSARE1, PSATEST1, PSATIME, PCPSARS1, PCPSADE1, PCDMDECN,
          `_DRNKWEK`) %>%
    mutate(year = "2015")
#
 BRFSS_2016 <- BRFSS_2016 %>%
   select(SEX, `_DRNKWEK`, TRNSGNDR, `_RFDRHV5`, PREGNANT, HADMAM,
          HOWLONG, HADPAP2, LASTPAP2, HPVTEST, HPLSTTST, HADHYST2,
          PROFEXAM, LENGEXAM, PCPSAAD2, PCPSADI1, PCPSARE1, PSATEST1,
          PSATIME, PCPSARS1, PCPSADE1, PCDMDECN, `_RFMAM2Y`, `_MAM5021`,
          `_RFPAP33`, `_RFPSA21`)%>%
    mutate(year = "2016")
#
 BRFSS_2017 <- BRFSS_2017 %>%
   select(SEX, `_DRNKWEK`, TRNSGNDR, `_RFDRHV5`, PREGNANT, PFPPRVN2,
          TYPCNTR7, NOBCUSE6)%>%
    mutate(year = "2017")
#
 BRFSS_2018 <- BRFSS_2018 %>%
  select(SEX1, `_DRNKWEK`,  TRNSGNDR, `_LLCPWT`)%>%
    mutate(year = "2018")
#
 BRFSS_2019 <- BRFSS_2019 %>%
   select(`_MICHD`, `_SEX`, EDUCA, `_BMI5CAT`, `_DRNKWK1`, EXERANY2, `_RACE`,
          `_AGE80`, `_SMOKER3`, TRNSGNDR, `_LLCPWT`)%>%
    mutate(year = "2019")

 BRFSS_2020 <- BRFSS_2020 %>%
    mutate(year = "2020") %>%
    rename(SEX = `_SEX`)
#
# # renamed SEX1 in 2018 to SEX like all the other years
 BRFSS_2018 <- BRFSS_2018 %>%
   rename(SEX = SEX1)
 BRFSS_2019 <- BRFSS_2019 %>%
   rename(SEX = `_SEX`)
#
 # renamed AVEDRNK3 to AVEDRNK2 in 2019 like all the other years
BRFSS_2019 <- BRFSS_2019 %>%
  rename(`_DRNKWEK` = `_DRNKWK1`)




# # combine all years into 1 dataset
BRFSS_all <- rbind(select(BRFSS_2015, SEX, TRNSGNDR, year),
                   select(BRFSS_2016, SEX, TRNSGNDR, year),
                   select(BRFSS_2017, SEX, TRNSGNDR, year),
                   select(BRFSS_2018, SEX, TRNSGNDR, year),
                   select(BRFSS_2019, SEX, TRNSGNDR, year),
                   select(BRFSS_2020, SEX, TRNSGNDR, year))

# save BRFSS_all dataset as a .csv file so every time i want to work with, I just read that file
# in instead of running all steps above
# write.csv(BRFSS_all, file = "brfss_all.csv")
```

```{r}
BRFSS_all <- read.csv("brfss_all.csv")
# add labels
BRFSS_all_clean <- BRFSS_all %>% 
   mutate(SEX = recode_factor(SEX, 
                              '1' = 'male',
                              '2' = 'female')) %>% 
   mutate(TRNSGNDR = recode_factor(TRNSGNDR,
                                   '1' = 'Yes, Transgender, male-to-female',
                                   '2' = 'Yes, Transgender, female to male',
                                   '3' = 'Yes, Transgender, gender nonconforming',
                                   '4' = 'No',
                                   '7' = NA_character_,
                                   '9' = NA_character_)) %>% 
   filter(TRNSGNDR %in% c("Yes, Transgender, male-to-female",
                          "Yes, Transgender, female to male")) %>% 
   droplevels()
# basic table exploring data
table(trans = BRFSS_all_clean$TRNSGNDR, sex = BRFSS_all_clean$SEX, year = BRFSS_all_clean$year)

# library(table1)
# 
# table1(~ year | SEX * TRNSGNDR, data = BRFSS_all_clean)

prop.table(table(trans = BRFSS_all_clean$TRNSGNDR, 
                 sex = BRFSS_all_clean$SEX),
           margin = 2)
```

In 2021, a Gallup poll found that .6% or 60 out of every 1000 Americans, reported that they identify as transgender (add cite: <https://news.gallup.com/poll/329708/lgbt-identification-rises-latest-estimate.aspx>). The percentage varied by age group, with the youngest cohort, Generation Z born 1997-2002, reporting 1.8% of survey participants identified as transgender compared to 1.2% in the previous cohort (Millenials) and just .3% in the oldest cohort born before 1946 (Traditionalists). Transgender people are... Although the terms sex and gender are used interchangeably by... Gender identity is distinct from sex and sexual orientation and consists of...

The BRFSS is...

Until 2015, the Behavioral Risk Factor Surveillance Survey (BRFSS) did not ask participants about gender identity. In 2015 the BRFSS added an optional question about gender identity to the survey and states could decide whether or not to include the gender identity question when conducting data collection. Although a gender identity question was added, the wording of the sex question did not make any distinction between sex and gender for several years. In 2015 to 2017, the interviewer selected the sex of the participant given the prompt: "Indicate sex of respondent." In 2018, the interviewer asked, "What is your sex? or What was your sex at birth? Was it..." and gave the possible responses of male and female. In 2019, the interviewer asked, "Are you male or female?". In 2020, the sex question was changed to, "What was your sex at birth? Was it male or female?"

In a recent paper [@howerton2021] the authors found that transgender participants did not give a response to the sex question that was consistent with respect to their gender identity response. For example, some participants who identified as female-to-male transgender identity reported their sex as male and others reported their sex as female. Likewise, among those who identified as male-to-female transgender identity, some reported their sex to be male and others reported their sex to be female.

Scientific research on transgender healthcare is underdeveloped and the role of sex recorded at birth in health and healthcare remains unclear for transgender people [citations needed see alcohol literature]. However, sex recorded at birth is a risk factor for numerous types of cancer and other health conditions. In order to better understand the role of sex recorded at birth

+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------------+
| Question id | Question                                                                                                                                                                                     | Asked of... | Year(s)                |
+=============+==============================================================================================================================================================================================+=============+========================+
| \_DRNKWK1   |                                                                                                                                                                                              | computed    | 2015                   |
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------------+
| \_RFBING5   | Binge drinkers (males having five or more drinks on one occasion, females having four or more drinks on one occasion)                                                                        | computed    | 2015                   |
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------------+
| \_RFDRHV5   | Heavy drinkers (adult men having more than 14 drinks per week and adult women having more than 7 drinks per week)                                                                            | computed    | 2015, 2016, 2017       |
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------------+
| DIABETE3    | (Ever told) you have diabetes (If "Yes" and respondent is female, ask "Was this only when you were pregnant?". If Respondent says pre-diabetes or borderline diabetes, use response code 4.) | female      | 2015 2018              |
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------------+
| PREGNANT    | To your knowledge, are you now pregnant?                                                                                                                                                     | female      | 2015, 2016, 2017, 2018 |
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------------+
| HADMAM      | A mammogram is an x-ray of each breast to look for breast cancer. Have you ever had a mammogram?                                                                                             | female      | 2015, 2016, 2018       |
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------------+
| HOWLONG     | How long has it been since you had your last mammogram?                                                                                                                                      | female      | 2015, 2016, 2018       |
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------------+
| HADPAP2     | A Pap test is a test for cancer of the cervix. Have you ever had a Pap test?                                                                                                                 | female      | 2015, 2016, 2018       |
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------------+
| LASTPAP2    | How long has it been since you had your last Pap test?                                                                                                                                       | female      | 2015, 2016, 2018       |
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------------+
| HPVTEST     | An HPV test is sometimes given with the Pap test for cervical cancer screening. Have you ever had an HPV test?                                                                               | female      | 2015, 2016, 2018       |
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------------+
| HPLSTTST    | How long has it been since you had your last HPV test?                                                                                                                                       | female      | 2015, 2016, 2018       |
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------------+
| HADHYST2    | Have you had a hysterectomy? (A hysterectomy is an operation to remove the uterus (womb).)                                                                                                   | female      | 2015, 2016, 2018       |
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------------+
| PROFEXAM    | A clinical breast exam is when a doctor, nurse, or other health professional feels the breast for lumps. Have you ever had a clinical breast exam?                                           | female      | 2015, 2016             |
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------------+
| LENGEXAM    | How long has it been since your last breast exam?                                                                                                                                            | female      | 2015, 2016             |
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------------+
| PCPSAAD2    | Has a doctor, nurse, or other health professional EVER talked with you about the advantages of the PSA test?                                                                                 | male        | 2015, 2016             |
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------------+

---------------------+-------------+------------------+ \| PCPSAAD3 \| Has a doctor, nurse, or other health professional EVER talked with you about the advantages of the PSA test? \| male \| 2018 \| +-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+ \| PCPSADI1 \| Has a doctor, nurse, or other health professional EVER talked with you about the disadvantages of the PSA test? \| male \| 2015, 2016, 2018 \| +-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+ \| PCPSARE1 \| Has a doctor, nurse, or other health professional EVER recommended that you have a PSA test? \| male \| 2015, 2016, 2018 \| +-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+ \| PSATEST1 \| Have you EVER HAD a PSA test? \| male \| 2015, 2016, 2018 \| +-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+ \| PSATIME \| How long has it been since you had your last PSA test? \| male \| 2015, 2016, 2018 \| +-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+ \| PCPSARS1 \| What was the MAIN reason you had this PSA test -- was it ...? \| male \| 2015, 2016, 2018 \| +-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+ \| PCPSADE1 \| Which of the following best describes the decision to have the PSA test done? \| male \| 2015, 2016 \| +-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+ \| PCDMDECN \| Who made the decision with you? (Mark all that apply) \| male \| 2015, 2016 \| +-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+ \| \_RFMAM2Y  \| Women respondents aged 40+ who have had a mammogram in the past two years \| computed \| 2016 \| +-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+ \| \_MAM5021  \| Women respondents aged 50-74 who have had a mammogram in the past two years \| computed \| 2016 \| +-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+ \| \_RFPAP33  \| Women respondents aged 21-65 who have had a pap test in the past three years \| computed \| 2016 \| +-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+ \| \_RFPSA21  \| Male respondents aged 40+ who have had a PSA test in the past 2 years \| computed \| 2016 \| +-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+ \| PFPPRVN2  \| Did you or your partner do anything the last time you had sex to keep you from getting pregnant? \| female \| 2017 \| +-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+ \| TYPCNTR7  \| What did you or your partner do the last time you had sex to keep you from getting pregnant? \| female \| 2017 \| +-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+ \| NOBCUSE6  \| What was your main reason for not doing anything the last time you had sex to keep you from getting pregnant? \| female \| 2017 \| +-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+ ---------------------+-------------+------------------+ \| CNCRTYP1  \| With your most recent diagnoses of cancer, what type of cancer was it? \| female, male \| 2018 \| +-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
