---
title: "Measurement of sex and gender identity in BRFSS"
author: "Jenine K Harris, Odalis Hernandez"
date: "9/15/2021"
output: word_document
bibliography: references.bib
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = FALSE, message = FALSE,
                      warning = FALSE)
```

```{r}
# TO CREATE ANALYTIC DATA SET, UNCOMMENT AND RUN THIS CODE CHUNK
# bring in files from online BRFSS 2015 to 2018 data
library(haven)
# #
# #
# # # 2015 data
  temp <- tempfile(fileext = ".zip")
  download.file(url  = "https://www.cdc.gov/brfss/annual_data/2015/files/LLCP2015XPT.zip", destfile = temp)
  BRFSS_2015 <- read_xpt(file = temp)
# #
# # 2016 data
 temp <- tempfile(fileext = ".zip")
 download.file(url  = "https://www.cdc.gov/brfss/annual_data/2016/files/LLCP2016XPT.zip", destfile = temp)
 BRFSS_2016 <- read_xpt(file = temp)
#
# # 2017 data
 temp <- tempfile(fileext = ".zip")
 download.file(url  = "https://www.cdc.gov/brfss/annual_data/2017/files/LLCP2017XPT.zip", destfile = temp)
 BRFSS_2017 <- read_xpt(file = temp)
#
# 2018 data
temp <- tempfile(fileext = ".zip")
download.file(url  = "https://www.cdc.gov/brfss/annual_data/2018/files/LLCP2018XPT.zip", destfile = temp)
BRFSS_2018 <- read_xpt(file = temp)

# 2019 data
temp <- tempfile(fileext = ".zip")
download.file(url  = "https://www.cdc.gov/brfss/annual_data/2019/files/LLCP2019XPT.zip", destfile = temp)
BRFSS_2019 <- read_xpt(file = temp)

# 2020 data
temp <- tempfile(fileext = ".zip")
download.file(url  = "https://www.cdc.gov/brfss/annual_data/2020/files/LLCP2020XPT.zip", destfile = temp)
BRFSS_2020 <- read_xpt(file = temp)
```

```{r}
#
# # select only the variables to use in the analysis (e.g. changes 300+ variables to only needed for analysis)

#
library(tidyverse)
 BRFSS_2015 <- BRFSS_2015 %>%
   select(SEX, TRNSGNDR, `_RFBING5`, `_RFDRHV5`, DIABETE3,
          PREGNANT, HADMAM, HOWLONG, HADPAP2, LASTPAP2, HPVTEST,
          HPLSTTST, HADHYST2, PROFEXAM, LENGEXAM, PCPSAAD2, PCPSADI1,
          PCPSARE1, PSATEST1, PSATIME, PCPSARS1, PCPSADE1, PCDMDECN,
          `_DRNKWEK`) %>%
    mutate(year = "2015")
#
 BRFSS_2016 <- BRFSS_2016 %>%
   select(SEX, `_DRNKWEK`, TRNSGNDR, `_RFDRHV5`, PREGNANT, HADMAM,
          HOWLONG, HADPAP2, LASTPAP2, HPVTEST, HPLSTTST, HADHYST2,
          PROFEXAM, LENGEXAM, PCPSAAD2, PCPSADI1, PCPSARE1, PSATEST1,
          PSATIME, PCPSARS1, PCPSADE1, PCDMDECN, `_RFMAM2Y`, `_MAM5021`,
          `_RFPAP33`, `_RFPSA21`)%>%
    mutate(year = "2016")
#
 BRFSS_2017 <- BRFSS_2017 %>%
   select(SEX, `_DRNKWEK`, TRNSGNDR, `_RFDRHV5`, PREGNANT, PFPPRVN2,
          TYPCNTR7, NOBCUSE6)%>%
    mutate(year = "2017")
#
 BRFSS_2018 <- BRFSS_2018 %>%
  select(SEX1, `_DRNKWEK`,  TRNSGNDR, `_LLCPWT`)%>%
    mutate(year = "2018")
#
 BRFSS_2019 <- BRFSS_2019 %>%
   select(`_MICHD`, `_SEX`, EDUCA, `_BMI5CAT`, `_DRNKWK1`, EXERANY2, `_RACE`,
          `_AGE80`, `_SMOKER3`, TRNSGNDR, `_LLCPWT`)%>%
    mutate(year = "2019")

 BRFSS_2020 <- BRFSS_2020 %>%
    mutate(year = "2020") %>%
    rename(SEX = `_SEX`)
#
# # renamed SEX1 in 2018 to SEX like all the other years
 BRFSS_2018 <- BRFSS_2018 %>%
   rename(SEX = SEX1)
 BRFSS_2019 <- BRFSS_2019 %>%
   rename(SEX = `_SEX`)
#
 # renamed AVEDRNK3 to AVEDRNK2 in 2019 like all the other years
BRFSS_2019 <- BRFSS_2019 %>%
  rename(`_DRNKWEK` = `_DRNKWK1`)




# # combine all years into 1 dataset
BRFSS_all <- rbind(select(BRFSS_2015, SEX, TRNSGNDR, year),
                   select(BRFSS_2016, SEX, TRNSGNDR, year),
                   select(BRFSS_2017, SEX, TRNSGNDR, year),
                   select(BRFSS_2018, SEX, TRNSGNDR, year),
                   select(BRFSS_2019, SEX, TRNSGNDR, year),
                   select(BRFSS_2020, SEX, TRNSGNDR, year))

# save BRFSS_all dataset as a .csv file so every time i want to work with, I just read that file
# in instead of running all steps above
# write.csv(BRFSS_all, file = "brfss_all.csv")
```

```{r}
#BRFSS_all <- read.csv("brfss_all.csv")
# add labels
BRFSS_all_clean <- BRFSS_all %>% 
   mutate(SEX = recode_factor(SEX, 
                              '1' = 'male',
                              '2' = 'female')) %>% 
   mutate(TRNSGNDR = recode_factor(TRNSGNDR,
                                   '1' = 'Trans male-to-female',
                                   '2' = 'Trans female-to-male',
                                   '3' = 'Trans gender nonconforming',
                                   '4' = 'No',
                                   '7' = NA_character_,
                                   '9' = NA_character_)) %>% 
   filter(TRNSGNDR %in% c("Trans male-to-female",
                          "Trans female-to-male")) %>% 
  mutate(yearLong = recode_factor(year,
                              '2015' = '2015: Indicate sex of respondent.',
                              '2016' = '2016: Indicate sex of respondent.',
                              '2017' = '2017: Indicate sex of respondent.',
                              '2018' = '2018: What is
your sex? or What was your sex at birth? Was it...',
                              '2019' = '2019: Are you male or female?',
'2020' = '2020: What was your sex at birth? Was it male or female?')) %>% 
  drop_na() %>% 
   droplevels()
```

In 2021, a Gallup poll found that .6% or 60 out of every 1000 Americans,
reported that they identify as transgender [@lgbtgallup]. The percentage
varied by age group, with the youngest cohort, Generation Z born
1997-2002, reporting 1.8% of survey participants identified as
transgender compared to 1.2% in the previous cohort (Millenials) and
just .3% in the cohort born before 1946 (Traditionalists). Transgender
people are individuals whose biological sex, or sex assigned at birth,
differs from their gender identity .(add cite:
<https://www.teachingtransgender.org/terminology-sheet-free-resource/>)
Gender identity is distinct from sex and sexual orientation and consists
of how one identifies and conceptualizes onself as a gendered being
along a spectrum of male, female, or in some cases, neither. Gender
identity is often perceived outwardly through observable presentation
that reflect internal identity, such as through behavior or appearance.
Alternately, sexual orientation describes external physical, emotional,
and sexual attraction toward others. add cite:
(<https://www.teachingtransgender.org/terminology-sheet-free-resource/>).
It is important to note that sexual orientation is independent of sex
and gender and not synonymous.

The Behavioral Risk Factor Surveillance Survey (BRFSS) collects data on
risk behaviors, health conditions, and the use of health-related
services from U.S residents. Respondent data is collected each month
over telephone to be aggregated and assessed by the CDC. It is
considered the world's largest health survey system, collecting data
from over 400,000 individuals yearly [@brfss].

Until 2015, the Behavioral Risk Factor Surveillance Survey (BRFSS) did
not ask participants about gender identity. In 2015 the BRFSS added an
optional question about gender identity to the survey and states could
decide whether or not to include the gender identity question when
conducting data collection. Although a gender identity question was
added, the wording of the sex question did not make any distinction
between sex and gender for several years. In 2015 to 2017, the
interviewer selected the sex of the participant given the prompt:
"Indicate sex of respondent." In 2018, the interviewer asked, "What is
your sex? or What was your sex at birth? Was it..." and gave the
possible responses of male and female. In 2019, the interviewer asked,
"Are you male or female?". The sex question did not change substantially
until the 2020 survey administration, where it was changed to multiple
prompts depending on the mode of data collection (cell phone, etc.),
"What was your sex at birth? Was it male or female?"

In a recent paper [@howerton2021] Howerton and Harris found that
transgender participants did not give a response to the sex question
that was consistent with respect to their gender identity response. For
example, some participants who identified as female-to-male transgender
identity reported their sex as male and others reported their sex as
female. Likewise, among those who identified as male-to-female
transgender identity, some reported their sex to be male and others
reported their sex to be female.

Scientific research on transgender healthcare is underdeveloped and the
role of sex recorded at birth in health and healthcare remains unclear
for transgender people [citations needed see alcohol literature and
screening literature]. However, sex recorded at birth is a risk factor
for numerous types of cancer and other health outcomes. In order to
better understand the role of sex recorded at birth in health and health
outcomes for transgender people, it may be useful to understand how to
best measure sex recorded at birth. This paper examines the ways sex was
measured in the BRFSS from 2015 (the first year of the gender identity
question) to the most recently released year of data in 2020.

### Methods

#### Data

We imported the BRFSS data sets from 2015 through 2020 directly from the
Centers for Disease Control and Prevention website. We used R
statistical software version 4.1.1 [@rstats], the RStudio (version
1.4.1717) interactive development environment [@rstudio], and a function
from version 2.4.3 of the haven package [@haven] to import the data.

```{r}
# # basic table exploring data
# table(trans = BRFSS_all_clean$TRNSGNDR, sex = BRFSS_all_clean$SEX, year = BRFSS_all_clean$year)
# 
# # library(table1)
# # 
# # table1(~ year | SEX * TRNSGNDR, data = BRFSS_all_clean)
# 
# prop.table(table(trans = BRFSS_all_clean$TRNSGNDR, 
#                  sex = BRFSS_all_clean$SEX),
#            margin = 2)

BRFSS_all_clean %>% 
  drop_na() %>% 
  group_by(TRNSGNDR, SEX, yearLong) %>% 
  count() %>% 
  group_by(TRNSGNDR, yearLong) %>% 
  mutate(percYear = 100 * n / sum(n)) %>% 
  ggplot(aes(x = TRNSGNDR, fill = SEX, y = percYear)) +
  geom_col(position = "dodge") +
  #coord_flip() +
  scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 10)) +
  facet_wrap(~yearLong, 
             labeller = label_wrap_gen(20)) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  labs(x = "Gender identity", y = "Percent in each sex category",
       fill = "Sex category")  

```

+---+--------------------------------------------------------+---+----+
| Q | Question                                               | A | Y  |
| u |                                                        | s | ea |
| e |                                                        | k | r( |
| s |                                                        | e | s) |
| t |                                                        | d |    |
| i |                                                        | o |    |
| o |                                                        | f |    |
| n |                                                        | . |    |
| i |                                                        | . |    |
| d |                                                        | . |    |
+===+========================================================+===+====+
| \ |                                                        | c | 20 |
| \ |                                                        | o | 15 |
| \ |                                                        | m |    |
| _ |                                                        | p |    |
| D |                                                        | u |    |
| R |                                                        | t |    |
| N |                                                        | e |    |
| K |                                                        | d |    |
| W |                                                        |   |    |
| K |                                                        |   |    |
| 1 |                                                        |   |    |
+---+--------------------------------------------------------+---+----+
| \ | Binge drinkers (males having five or more drinks on    | c | 20 |
| \ | one occasion, females having four or more drinks on    | o | 15 |
| \ | one occasion)                                          | m |    |
| _ |                                                        | p |    |
| R |                                                        | u |    |
| F |                                                        | t |    |
| B |                                                        | e |    |
| I |                                                        | d |    |
| N |                                                        |   |    |
| G |                                                        |   |    |
| 5 |                                                        |   |    |
+---+--------------------------------------------------------+---+----+
| \ | Heavy drinkers (adult men having more than 14 drinks   | c | 2  |
| \ | per week and adult women having more than 7 drinks per | o | 01 |
| \ | week)                                                  | m | 5, |
| _ |                                                        | p | 2  |
| R |                                                        | u | 01 |
| F |                                                        | t | 6, |
| D |                                                        | e | 20 |
| R |                                                        | d | 17 |
| H |                                                        |   |    |
| V |                                                        |   |    |
| 5 |                                                        |   |    |
+---+--------------------------------------------------------+---+----+
| \ | Heavy drinkers (adult men having more than 14 drinks   | c | 20 |
| \ | per week and adult women having more than 7 drinks per | o | 18 |
| \ | week)                                                  | m |    |
| _ |                                                        | p |    |
| R |                                                        | u |    |
| F |                                                        | t |    |
| D |                                                        | e |    |
| R |                                                        | d |    |
| H |                                                        |   |    |
| V |                                                        |   |    |
| 6 |                                                        |   |    |
+---+--------------------------------------------------------+---+----+
| D | (Ever told) you have diabetes (If "Yes" and respondent | f | 20 |
| I | is female, ask "Was this only when you were            | e | 15 |
| A | pregnant?". If Respondent says pre-diabetes or         | m | 20 |
| B | borderline diabetes, use response code 4.)             | a | 18 |
| E |                                                        | l |    |
| T |                                                        | e |    |
| E |                                                        |   |    |
| 3 |                                                        |   |    |
+---+--------------------------------------------------------+---+----+
| P | To your knowledge, are you now pregnant?               | f | 2  |
| R |                                                        | e | 01 |
| E |                                                        | m | 5, |
| G |                                                        | a | 2  |
| N |                                                        | l | 01 |
| A |                                                        | e | 6, |
| N |                                                        |   | 2  |
| T |                                                        |   | 01 |
|   |                                                        |   | 7, |
|   |                                                        |   | 20 |
|   |                                                        |   | 18 |
+---+--------------------------------------------------------+---+----+
| H | A mammogram is an x-ray of each breast to look for     | f | 2  |
| A | breast cancer. Have you ever had a mammogram?          | e | 01 |
| D |                                                        | m | 5, |
| M |                                                        | a | 2  |
| A |                                                        | l | 01 |
| M |                                                        | e | 6, |
|   |                                                        |   | 20 |
|   |                                                        |   | 18 |
+---+--------------------------------------------------------+---+----+
| H | How long has it been since you had your last           | f | 2  |
| O | mammogram?                                             | e | 01 |
| W |                                                        | m | 5, |
| L |                                                        | a | 2  |
| O |                                                        | l | 01 |
| N |                                                        | e | 6, |
| G |                                                        |   | 20 |
|   |                                                        |   | 18 |
+---+--------------------------------------------------------+---+----+
| H | A Pap test is a test for cancer of the cervix. Have    | f | 2  |
| A | you ever had a Pap test?                               | e | 01 |
| D |                                                        | m | 5, |
| P |                                                        | a | 2  |
| A |                                                        | l | 01 |
| P |                                                        | e | 6, |
| 2 |                                                        |   | 20 |
|   |                                                        |   | 18 |
+---+--------------------------------------------------------+---+----+
| L | How long has it been since you had your last Pap test? | f | 2  |
| A |                                                        | e | 01 |
| S |                                                        | m | 5, |
| T |                                                        | a | 2  |
| P |                                                        | l | 01 |
| A |                                                        | e | 6, |
| P |                                                        |   | 20 |
| 2 |                                                        |   | 18 |
+---+--------------------------------------------------------+---+----+
| H | An HPV test is sometimes given with the Pap test for   | f | 2  |
| P | cervical cancer screening. Have you ever had an HPV    | e | 01 |
| V | test?                                                  | m | 5, |
| T |                                                        | a | 2  |
| E |                                                        | l | 01 |
| S |                                                        | e | 6, |
| T |                                                        |   | 20 |
|   |                                                        |   | 18 |
+---+--------------------------------------------------------+---+----+
| H | How long has it been since you had your last HPV test? | f | 2  |
| P |                                                        | e | 01 |
| L |                                                        | m | 5, |
| S |                                                        | a | 2  |
| T |                                                        | l | 01 |
| T |                                                        | e | 6, |
| S |                                                        |   | 20 |
| T |                                                        |   | 18 |
+---+--------------------------------------------------------+---+----+
| H | Have you had a hysterectomy? (A hysterectomy is an     | f | 2  |
| A | operation to remove the uterus (womb).)                | e | 01 |
| D |                                                        | m | 5, |
| H |                                                        | a | 2  |
| Y |                                                        | l | 01 |
| S |                                                        | e | 6, |
| T |                                                        |   | 20 |
| 2 |                                                        |   | 18 |
+---+--------------------------------------------------------+---+----+
| P | A clinical breast exam is when a doctor, nurse, or     | f | 2  |
| R | other health professional feels the breast for lumps.  | e | 01 |
| O | Have you ever had a clinical breast exam?              | m | 5, |
| F |                                                        | a | 2  |
| E |                                                        | l | 01 |
| X |                                                        | e | 6, |
| A |                                                        |   | 20 |
| M |                                                        |   | 18 |
+---+--------------------------------------------------------+---+----+
| L | How long has it been since your last breast exam?      | f | 2  |
| E |                                                        | e | 01 |
| N |                                                        | m | 5, |
| G |                                                        | a | 2  |
| E |                                                        | l | 01 |
| X |                                                        | e | 6, |
| A |                                                        |   | 20 |
| M |                                                        |   | 18 |
+---+--------------------------------------------------------+---+----+
| P | Has a doctor, nurse, or other health professional EVER | m | 2  |
| C | talked with you about the advantages of the PSA test?  | a | 01 |
| P |                                                        | l | 5, |
| S |                                                        | e | 20 |
| A |                                                        |   | 16 |
| A |                                                        |   |    |
| D |                                                        |   |    |
| 2 |                                                        |   |    |
+---+--------------------------------------------------------+---+----+
| P | Has a doctor, nurse, or other health professional EVER | m | 20 |
| C | talked with you about the advantages of the PSA test?  | a | 18 |
| P |                                                        | l |    |
| S |                                                        | e |    |
| A |                                                        |   |    |
| A |                                                        |   |    |
| D |                                                        |   |    |
| 3 |                                                        |   |    |
+---+--------------------------------------------------------+---+----+
| P | Has a doctor, nurse, or other health professional EVER | m | 2  |
| C | talked with you about the disadvantages of the PSA     | a | 01 |
| P | test?                                                  | l | 5, |
| S |                                                        | e | 2  |
| A |                                                        |   | 01 |
| D |                                                        |   | 6, |
| I |                                                        |   | 20 |
| 1 |                                                        |   | 18 |
+---+--------------------------------------------------------+---+----+
| P | Has a doctor, nurse, or other health professional EVER | m | 2  |
| C | recommended that you have a PSA test?                  | a | 01 |
| P |                                                        | l | 5, |
| S |                                                        | e | 2  |
| A |                                                        |   | 01 |
| R |                                                        |   | 6, |
| E |                                                        |   | 20 |
| 1 |                                                        |   | 18 |
+---+--------------------------------------------------------+---+----+
| P | Have you EVER HAD a PSA test?                          | m | 2  |
| S |                                                        | a | 01 |
| A |                                                        | l | 5, |
| T |                                                        | e | 2  |
| E |                                                        |   | 01 |
| S |                                                        |   | 6, |
| T |                                                        |   | 20 |
| 1 |                                                        |   | 18 |
+---+--------------------------------------------------------+---+----+
| P | How long has it been since you had your last PSA test? | m | 2  |
| S |                                                        | a | 01 |
| A |                                                        | l | 5, |
| T |                                                        | e | 2  |
| I |                                                        |   | 01 |
| M |                                                        |   | 6, |
| E |                                                        |   | 20 |
|   |                                                        |   | 18 |
+---+--------------------------------------------------------+---+----+
| P | What was the MAIN reason you had this PSA test -- was  | m | 2  |
| C | it ...?                                                | a | 01 |
| P |                                                        | l | 5, |
| S |                                                        | e | 2  |
| A |                                                        |   | 01 |
| R |                                                        |   | 6, |
| S |                                                        |   | 20 |
| 1 |                                                        |   | 18 |
+---+--------------------------------------------------------+---+----+
| P | Which of the following best describes the decision to  | m | 2  |
| C | have the PSA test done?                                | a | 01 |
| P |                                                        | l | 5, |
| S |                                                        | e | 2  |
| A |                                                        |   | 01 |
| D |                                                        |   | 6, |
| E |                                                        |   | 20 |
| 1 |                                                        |   | 18 |
+---+--------------------------------------------------------+---+----+
| P | Who made the decision with you? (Mark all that apply)  | m | 2  |
| C |                                                        | a | 01 |
| D |                                                        | l | 5, |
| M |                                                        | e | 20 |
| D |                                                        |   | 16 |
| E |                                                        |   |    |
| C |                                                        |   |    |
| N |                                                        |   |    |
+---+--------------------------------------------------------+---+----+
| \ | Women respondents aged 40+ who have had a mammogram in | c | 20 |
| _ | the past two years                                     | o | 16 |
| R |                                                        | m |    |
| F |                                                        | p |    |
| M |                                                        | u |    |
| A |                                                        | t |    |
| M |                                                        | e |    |
| 2 |                                                        | d |    |
| Y |                                                        |   |    |
+---+--------------------------------------------------------+---+----+
| \ | Women respondents aged 50-74 who have had a mammogram  | c | 20 |
| _ | in the past two years                                  | o | 16 |
| M |                                                        | m |    |
| A |                                                        | p |    |
| M |                                                        | u |    |
| 5 |                                                        | t |    |
| 0 |                                                        | e |    |
| 2 |                                                        | d |    |
| 1 |                                                        |   |    |
+---+--------------------------------------------------------+---+----+
| \ | Women respondents aged 21-65 who have had a pap test   | c | 20 |
| _ | in the past three years                                | o | 16 |
| R |                                                        | m |    |
| F |                                                        | p |    |
| P |                                                        | u |    |
| A |                                                        | t |    |
| P |                                                        | e |    |
| 3 |                                                        | d |    |
| 3 |                                                        |   |    |
+---+--------------------------------------------------------+---+----+
| \ | Male respondents aged 40+ who have had a PSA test in   | c | 20 |
| _ | the past 2 years                                       | o | 16 |
| R |                                                        | m |    |
| F |                                                        | p |    |
| P |                                                        | u |    |
| S |                                                        | t |    |
| A |                                                        | e |    |
| 2 |                                                        | d |    |
| 1 |                                                        |   |    |
+---+--------------------------------------------------------+---+----+
| P | Did you or your partner do anything the last time you  | f | 20 |
| F | had sex to keep you from getting pregnant?             | e | 17 |
| P |                                                        | m |    |
| P |                                                        | a |    |
| R |                                                        | l |    |
| V |                                                        | e |    |
| N |                                                        |   |    |
| 2 |                                                        |   |    |
+---+--------------------------------------------------------+---+----+
| T | What did you or your partner do the last time you had  | f | 20 |
| Y | sex to keep you from getting pregnant?                 | e | 17 |
| P |                                                        | m |    |
| C |                                                        | a |    |
| N |                                                        | l |    |
| T |                                                        | e |    |
| R |                                                        |   |    |
| 7 |                                                        |   |    |
+---+--------------------------------------------------------+---+----+
| N | What was your main reason for not doing anything the   | f | 20 |
| O | last time you had sex to keep you from getting         | e | 17 |
| B | pregnant?                                              | m |    |
| C |                                                        | a |    |
| U |                                                        | l |    |
| S |                                                        | e |    |
| E |                                                        |   |    |
| 6 |                                                        |   |    |
+---+--------------------------------------------------------+---+----+
| C | With your most recent diagnoses of cancer, what type   | f | 20 |
| N | of cancer was it?                                      | e | 18 |
| C |                                                        | m |    |
| R |                                                        | a |    |
| T |                                                        | l |    |
| Y |                                                        | e |    |
| P |                                                        | , |    |
| 1 |                                                        | m |    |
|   |                                                        | a |    |
|   |                                                        | l |    |
|   |                                                        | e |    |
+---+--------------------------------------------------------+---+----+
| C | What type of doctor provides the majority of your      | f | 20 |
| S | health care?                                           | e | 18 |
| R |                                                        | m |    |
| V |                                                        | a |    |
| D |                                                        | l |    |
| O |                                                        | e |    |
| C |                                                        |   |    |
| 1 |                                                        |   |    |
+---+--------------------------------------------------------+---+----+
| S | Which of the following best represents how you think   | m | 20 |
| O | of yourself?                                           | a | 18 |
| M |                                                        | l |    |
| A |                                                        | e |    |
| L |                                                        |   |    |
| E |                                                        |   |    |
+---+--------------------------------------------------------+---+----+
| S | Which of the following best represents how you think   | f | 20 |
| O | of yourself?                                           | e | 18 |
| F |                                                        | m |    |
| E |                                                        | a |    |
| M |                                                        | l |    |
| A |                                                        | e |    |
| L |                                                        |   |    |
| E |                                                        |   |    |
+---+--------------------------------------------------------+---+----+
| \ | Women respondents aged 40+ who have had a mammogram in | f | 20 |
| _ | the past two years?                                    | e | 18 |
| R |                                                        | m |    |
| F |                                                        | a |    |
| M |                                                        | l |    |
| A |                                                        | e |    |
| M |                                                        |   |    |
| 2 |                                                        |   |    |
| 1 |                                                        |   |    |
+---+--------------------------------------------------------+---+----+
| \ | Women respondents aged 50-74 who have had a mammogram  | f | 20 |
| _ | in the past two years?                                 | e | 18 |
| M |                                                        | m |    |
| A |                                                        | a |    |
| M |                                                        | l |    |
| 5 |                                                        | e |    |
| 0 |                                                        |   |    |
| 2 |                                                        |   |    |
| 2 |                                                        |   |    |
+---+--------------------------------------------------------+---+----+
| \ | Women respondents aged 50-74 who have had a pap test   | f | 20 |
| _ | in the past two years?                                 | e | 18 |
| R |                                                        | m |    |
| F |                                                        | a |    |
| P |                                                        | l |    |
| A |                                                        | e |    |
| P |                                                        |   |    |
| 3 |                                                        |   |    |
| 4 |                                                        |   |    |
+---+--------------------------------------------------------+---+----+
| \ | Male respondents aged 40+ who have had a PSA test in   | m | 20 |
| _ | the past two years?                                    | a | 18 |
| R |                                                        | l |    |
| F |                                                        | e |    |
| P |                                                        |   |    |
| S |                                                        |   |    |
| A |                                                        |   |    |
| 2 |                                                        |   |    |
| 2 |                                                        |   |    |
+---+--------------------------------------------------------+---+----+
