---
title: "Measurement of sex and gender identity in BRFSS"
author: "Jenine K Harris, Odalis Hernandez"
date: "9/15/2021"
output: word_document
bibliography: references.bib
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = FALSE, message = FALSE,
                      warning = FALSE)
```

```{r eval = FALSE}
# # remove the eval = FALSE to run
# # TO CREATE ANALYTIC DATA SET, UNCOMMENT AND RUN THIS CODE CHUNK
# # bring in files from online BRFSS 2015 to 2018 data
# library(haven)
# # #
# # #
# # # # 2015 data
#   temp <- tempfile(fileext = ".zip")
#   download.file(url  = "https://www.cdc.gov/brfss/annual_data/2015/files/LLCP2015XPT.zip", destfile = temp)
#   BRFSS_2015 <- read_xpt(file = temp)
# # #
# # # 2016 data
#  temp <- tempfile(fileext = ".zip")
#  download.file(url  = "https://www.cdc.gov/brfss/annual_data/2016/files/LLCP2016XPT.zip", destfile = temp)
#  BRFSS_2016 <- read_xpt(file = temp)
# #
# # # 2017 data
#  temp <- tempfile(fileext = ".zip")
#  download.file(url  = "https://www.cdc.gov/brfss/annual_data/2017/files/LLCP2017XPT.zip", destfile = temp)
#  BRFSS_2017 <- read_xpt(file = temp)
# #
# # 2018 data
# temp <- tempfile(fileext = ".zip")
# download.file(url  = "https://www.cdc.gov/brfss/annual_data/2018/files/LLCP2018XPT.zip", destfile = temp)
# BRFSS_2018 <- read_xpt(file = temp)
# 
# # 2019 data
# temp <- tempfile(fileext = ".zip")
# download.file(url  = "https://www.cdc.gov/brfss/annual_data/2019/files/LLCP2019XPT.zip", destfile = temp)
# BRFSS_2019 <- read_xpt(file = temp)
# 
# # 2020 data
# temp <- tempfile(fileext = ".zip")
# download.file(url  = "https://www.cdc.gov/brfss/annual_data/2020/files/LLCP2020XPT.zip", destfile = temp)
# BRFSS_2020 <- read_xpt(file = temp)
```

```{r eval = FALSE}
# remove the eval = FALSE to run
# # select only the variables to use in the analysis (e.g. changes 300+ variables to only needed for analysis)

#
library(tidyverse)
#  BRFSS_2015 <- BRFSS_2015 %>%
#    select(SEX, TRNSGNDR, `_RFBING5`, `_RFDRHV5`, DIABETE3,
#           PREGNANT, HADMAM, HOWLONG, HADPAP2, LASTPAP2, HPVTEST,
#           HPLSTTST, HADHYST2, PROFEXAM, LENGEXAM, PCPSAAD2, PCPSADI1,
#           PCPSARE1, PSATEST1, PSATIME, PCPSARS1, PCPSADE1, PCDMDECN,
#           `_DRNKWEK`) %>%
#     mutate(year = "2015")
# #
#  BRFSS_2016 <- BRFSS_2016 %>%
#    select(SEX, `_DRNKWEK`, TRNSGNDR, `_RFDRHV5`, PREGNANT, HADMAM,
#           HOWLONG, HADPAP2, LASTPAP2, HPVTEST, HPLSTTST, HADHYST2,
#           PROFEXAM, LENGEXAM, PCPSAAD2, PCPSADI1, PCPSARE1, PSATEST1,
#           PSATIME, PCPSARS1, PCPSADE1, PCDMDECN, `_RFMAM2Y`, `_MAM5021`,
#           `_RFPAP33`, `_RFPSA21`)%>%
#     mutate(year = "2016")
# #
#  BRFSS_2017 <- BRFSS_2017 %>%
#    select(SEX, `_DRNKWEK`, TRNSGNDR, `_RFDRHV5`, PREGNANT, PFPPRVN2,
#           TYPCNTR7, NOBCUSE6)%>%
#     mutate(year = "2017")
# #
#  BRFSS_2018 <- BRFSS_2018 %>%
#   select(SEX1, `_DRNKWEK`,  TRNSGNDR, `_LLCPWT`)%>%
#     mutate(year = "2018")
# #
#  BRFSS_2019 <- BRFSS_2019 %>%
#    select(`_MICHD`, `_SEX`, EDUCA, `_BMI5CAT`, `_DRNKWK1`, EXERANY2, `_RACE`,
#           `_AGE80`, `_SMOKER3`, TRNSGNDR, `_LLCPWT`)%>%
#     mutate(year = "2019")
# 
#  BRFSS_2020 <- BRFSS_2020 %>%
#     mutate(year = "2020") %>%
#     rename(SEX = `_SEX`)
# #
# # # renamed SEX1 in 2018 to SEX like all the other years
#  BRFSS_2018 <- BRFSS_2018 %>%
#    rename(SEX = SEX1)
#  BRFSS_2019 <- BRFSS_2019 %>%
#    rename(SEX = `_SEX`)
# #
#  # renamed AVEDRNK3 to AVEDRNK2 in 2019 like all the other years
# BRFSS_2019 <- BRFSS_2019 %>%
#   rename(`_DRNKWEK` = `_DRNKWK1`)
# 
# 
# 
# 
# # # combine all years into 1 dataset
# BRFSS_all <- rbind(select(BRFSS_2015, SEX, TRNSGNDR, year),
#                    select(BRFSS_2016, SEX, TRNSGNDR, year),
#                    select(BRFSS_2017, SEX, TRNSGNDR, year),
#                    select(BRFSS_2018, SEX, TRNSGNDR, year),
#                    select(BRFSS_2019, SEX, TRNSGNDR, year),
#                    select(BRFSS_2020, SEX, TRNSGNDR, year))
# 
# # save BRFSS_all dataset as a .csv file so every time i want to work with, I just read that file
# # in instead of running all steps above
# write.csv(BRFSS_all, file = "brfss_all.csv")
```

```{r}
BRFSS_all <- read.csv("brfss_all.csv")
# add labels
BRFSS_all_clean <- BRFSS_all %>% 
   mutate(SEX = recode_factor(SEX, 
                              '1' = 'male',
                              '2' = 'female')) %>% 
   mutate(TRNSGNDR = recode_factor(TRNSGNDR,
                                   '1' = 'Trans male-to-female',
                                   '2' = 'Trans female-to-male',
                                   '3' = 'Trans gender nonconforming',
                                   '4' = 'No',
                                   '7' = NA_character_,
                                   '9' = NA_character_)) %>% 
   filter(TRNSGNDR %in% c("Trans male-to-female",
                          "Trans female-to-male")) %>% 
  mutate(yearLong = recode_factor(year,
                              '2015' = '2015: Indicate sex of respondent.',
                              '2016' = '2016: Indicate sex of respondent.',
                              '2017' = '2017: Indicate sex of respondent.',
                              '2018' = '2018: What is
your sex? or What was your sex at birth? Was it...',
                              '2019' = '2019: Are you male or female?',
'2020' = '2020: What was your sex at birth? Was it male or female?')) %>% 
  drop_na() %>% 
   droplevels()
```

In 2021, a Gallup poll found that .6% or 60 out of every 1000 Americans,
reported that they identify as transgender [@lgbtgallup]. The percentage
varied by age group, with the cohort born in 1997-2002 reporting 1.8% of
survey participants identified as transgender compared to 1.2% in the
previous cohort and just .3% in the cohort born before 1946. Transgender
people are individuals whose biological sex, or sex recorded at birth,
differs from their gender identity [@galupo2016labels; @howerton2021].
Gender identity is how a person identifies and conceptualizes their own
gender and is different from biological sex and sexual orientation
[@galupo2016labels; @tate2020undoing]. Sex, sexual orientation, and
gender identity all have implications for health, health care, and
health outcomes [@legato2016consideration; @agenor2016sexual;
@gonzales2016comparison; @mcclure2022gender].

Conducting research to understand factors that influence health, health
care, and health outcomes is an important part of social science and
medical research. Given the roles of sex, sexual orientation, and gender
identity in health, many social science and medical surveys seek to
measure sex and a growing number also collect sexual orientation and
gender identity information. Until recently, measurement of gender
identity was unusual and measurement of sex was often done by an
interviewer recording the sex they thought the participant was based on
voice or other physical cues [@bauer2017transgender] or asking "what is
your sex?" and recording either male or female. In addition to these
inaccurate ways of recording sex [@bauer2017transgender], researchers
have frequently treated gender as a synonym for sex
[@krieger2003genders; @johnson2009better; @tate2013two].

For decades there have been calls to improve the measurement of sex,
gender, and sexual orientation and attempts to identify accurate ways of
measuring these characteristics [@krieger2003genders; @tate2013two]. In
the meantime, researchers are utilizing data sets that include less
accurate ways of measuring these concepts. For example, the Behavioral
Risk Factor Surveillance Survey (BRFSS) collects data on risk behaviors,
health conditions, and the use of health-related services from U.S
residents. Respondent data is collected each month over telephone to be
aggregated and assessed by the CDC. It is considered the world's largest
health survey system, collecting data from over 400,000 individuals
yearly [@brfss].

Until 2015, BRFSS did not ask participants about gender identity. In
2015 the BRFSS added an optional question about gender identity to the
survey and states could decide whether or not to include the gender
identity question when conducting data collection. The gender identity
question that was included in the BRFSS from 2015 until 2020 (the most
recent year of data released) was: "Do you consider yourself to be
transgender?" The follow-up question for those who answered yes was, "Do
you consider yourself to be 1. Male-to-female, 2. Female-to-male, or 3.
Gender nonconforming?"

Although a gender identity question was added, the wording of the sex
question did not make any distinction between sex recorded and gender
for several years. In 2015 to 2017, the interviewer selected the sex of
the participant given the prompt: "Indicate sex of respondent." In 2018,
the interviewer asked, "What is your sex? or What was your sex at birth?
Was it..." and gave the possible responses of male and female. In 2019,
the interviewer asked, "Are you male or female?". In 2019 there was a
separate "Sex at Birth Optional Module" that included the question,
"What was your sex at birth? Was it male or female?" In the 2020 survey
administration, the optional module sex at birth prompt was used as the
sex question, "What was your sex at birth? Was it male or female?"

In a recent paper [@howerton2021], Howerton and Harris found that BRFSS
participants from 2015 to 2019 who reported having the same transgender
gender identity had different answers for the sex question. For example,
some participants who identified as female-to-male transgender identity
reported their sex as male and others reported their sex as female.
Likewise, among those who identified as male-to-female transgender
identity, some reported their sex to be male and others reported their
sex to be female. This has likely led to numerous published scientific
papers examining sex differences but potentially including participants
whose sex recorded at birth was something other than their sex as
recorded in the BRFSS.

We sought to better understand examined the relationship between sex and
gender identity in the BRFSS data from 2015 (the first year of the
gender identity question) to the most recently released year of data in
2020. Specifically we answer three questions: (1) What is the
relationship between sex and gender identity responses for participants
over this six year period?; (2) How has the relationship between
measured sex and gender identity changed as the sex question changed
over this six year period?; and (3) Did the 2020 version of the question
that specified "at birth" result in different patterns of sex and gender
identity reporting compared to previous years?

### Methods

#### Data

We imported the BRFSS data sets from 2015 through 2020 directly from the
Centers for Disease Control and Prevention website. We used R
statistical software version 4.1.1 [@rstats], the RStudio (version
1.4.1717) interactive development environment [@rstudio], and a function
from version 2.4.3 of the haven package [@haven] to import the data.

```{r}
# # basic table exploring data
# table(trans = BRFSS_all_clean$TRNSGNDR, sex = BRFSS_all_clean$SEX, year = BRFSS_all_clean$year)
# 
# # library(table1)
# # 
# # table1(~ year | SEX * TRNSGNDR, data = BRFSS_all_clean)
# 
# prop.table(table(trans = BRFSS_all_clean$TRNSGNDR, 
#                  sex = BRFSS_all_clean$SEX),
#            margin = 2)

BRFSS_all_clean %>% 
  drop_na() %>% 
  group_by(TRNSGNDR, SEX, yearLong) %>% 
  count() %>% 
  group_by(TRNSGNDR, yearLong) %>% 
  mutate(percYear = 100 * n / sum(n)) %>% 
  ggplot(aes(x = TRNSGNDR, fill = SEX, y = percYear)) +
  geom_col(position = "dodge") +
  #coord_flip() +
  scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 10)) +
  facet_wrap(~yearLong, 
             labeller = label_wrap_gen(20)) +
  scale_fill_manual(values = c("gray60", "black")) +
  theme_minimal() +
  labs(x = "Gender identity", y = "Percent in each sex category",
       fill = "Sex category")  

```

```{r}
library(descr)

chisqtrans <- function(yr) {
  CrossTable(BRFSS_all_clean$SEX[BRFSS_all_clean$year == yr],
           BRFSS_all_clean$TRNSGNDR[BRFSS_all_clean$year == yr],
           chisq = TRUE,
           sresid = TRUE,
          prop.c = FALSE,
          prop.t = FALSE,
          prop.chisq = FALSE,
          expected = TRUE)
}

chisqtrans(2015)
chisqtrans(2016)
chisqtrans(2017)
chisqtrans(2018)
chisqtrans(2019)
chisqtrans(2020)

```

Scientific research on transgender healthcare is underdeveloped and the
role of sex recorded at birth in health and healthcare remains unclear
for transgender people [citations needed see alcohol literature and
screening literature]. However, sex recorded at birth is a risk factor
for numerous types of cancer and other health outcomes. In order to
better understand the role of sex recorded at birth in health and health
outcomes for transgender people, it may be useful to understand how to
best measure sex recorded at birth.

+------------+-----------------------------+------------+------------+
| Q u e s t  | Question                    | A s k e d  | Y ea r( s) |
| i o n i d  |                             | o f . . .  |            |
+============+=============================+============+============+
| \          |                             | c o m p u  | 20 15      |
| \          |                             | t e d      |            |
| \_ D R N K |                             |            |            |
| W K 1      |                             |            |            |
+------------+-----------------------------+------------+------------+
| \          | Binge drinkers (males       | c o m p u  | 20 15 , 20 |
| \          | having five or more drinks  | t e d      | 18         |
| \_ R F B I | on one occasion, females    |            |            |
| N G 5      | having four or more drinks  |            |            |
|            | on one occasion)            |            |            |
+------------+-----------------------------+------------+------------+
| \          | Heavy drinkers (adult men   | c o m p u  | 2 01 5, 2  |
| \          | having more than 14 drinks  | t e d      | 01 6, 20   |
| \_ R F D R | per week and adult women    |            | 17         |
| H V 5      | having more than 7 drinks   |            |            |
|            | per week)                   |            |            |
+------------+-----------------------------+------------+------------+
| \          | Heavy drinkers (adult men   | c o m p u  | 20 18      |
| \          | having more than 14 drinks  | t e d      |            |
| \_ R F D R | per week and adult women    |            |            |
| H V 6      | having more than 7 drinks   |            |            |
|            | per week)                   |            |            |
+------------+-----------------------------+------------+------------+

+-------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
+-------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
---------------+-------------+------------------+ \| DIABETE3 \| (Ever
told) you have diabetes (If "Yes" and respondent is female, ask "Was
this only when you were pregnant?". If Respondent says pre-diabetes or
borderline diabetes, use response code 4.) \| female \| 2015 2018 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+

\| PREGNANT \| To your knowledge, are you now pregnant? \| female \|
2015, 2016, 2017, 2018, 2019 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| HADMAM \| A mammogram is an x-ray of each breast to look for breast
cancer. Have you ever had a mammogram? \| female \| 2015, 2016, 2018,
2019 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| HOWLONG \| How long has it been since you had your last mammogram? \|
female \| 2015, 2016, 2018, 2019 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| HADPAP2 \| A Pap test is a test for cancer of the cervix. Have you
ever had a Pap test? \| female \| 2015, 2016, 2018, 2019 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| LASTPAP2 \| How long has it been since you had your last Pap test? \|
female \| 2015, 2016, 2018, 2019 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| HPVTEST \| An HPV test is sometimes given with the Pap test for
cervical cancer screening. Have you ever had an HPV test? \| female \|
2015, 2016, 2018, 2019 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| HPLSTTST \| How long has it been since you had your last HPV test? \|
female \| 2015, 2016, 2018, 2019 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| HADHYST2 \| Have you had a hysterectomy? (A hysterectomy is an
operation to remove the uterus (womb).) \| female \| 2015, 2016, 2018,
2019 \|

+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| PROFEXAM \| A clinical breast exam is when a doctor, nurse, or other
health professional feels the breast for lumps. Have you ever had a
clinical breast exam? \| female \| 2015, 2016, 2018 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| LENGEXAM \| How long has it been since your last breast exam? \|
female \| 2015, 2016, 2018 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| PCPSAAD2 \| Has a doctor, nurse, or other health professional EVER
talked with you about the advantages of the PSA test? \| male \| 2015,
2016 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+

+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| PCPSARE1 \| Has a doctor, nurse, or other health professional EVER
recommended that you have a PSA test? \| male \| 2015, 2016, 2018, 2019
\|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+

+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| PCDMDECN \| Who made the decision with you? (Mark all that apply) \|
male \| 2015, 2016 \|
+-------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

+-------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
---------------+-------------+------------------+ \| \_RFMAM2Y  \| Women
respondents aged 40+ who have had a mammogram in the past two years \|
computed \| 2016 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| \_MAM5021  \| Women respondents aged 50-74 who have had a mammogram
in the past two years \| computed \| 2016 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| \_RFPAP33  \| Women respondents aged 21-65 who have had a pap test in
the past three years \| computed \| 2016 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| \_RFPSA21  \| Male respondents aged 40+ who have had a PSA test in
the past 2 years \| computed \| 2016 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| PFPPRVN2  \| Did you or your partner do anything the last time you
had sex to keep you from getting pregnant? \| female \| 2017, 2019 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| TYPCNTR7  \| What did you or your partner do the last time you had
sex to keep you from getting pregnant? \| female \| 2017 \|
+-------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
---------------+-------------+------------------+ \| TYPCNTR8  \| What
did you or your partner do the last time you had sex to keep you from
getting pregnant? \| female \| 2019 \|
+-------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
---------------+-------------+------------------+ \| NOBCUSE6  \| What
was your main reason for not doing anything the last time you had sex to
keep you from getting pregnant? \| female \| 2017 \|
+-------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
---------------+-------------+------------------+ \| NOBCUSE7  \| What
was your main reason for not doing anything the last time you had sex to
keep you from getting pregnant? \| female \| 2019 \|
+-------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
---------------+-------------+------------------+
---------------------+-------------+------------------+ \| CNCRTYP1  \|
With your most recent diagnoses of cancer, what type of cancer was it?
\| female, male \| 2018, 2019 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| CSRVDOC1  \| What type of doctor provides the majority of your health
care? \| female \| 2018, 2019 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+

+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| \_RFMAM21  \| Women respondents aged 40+ who have had a mammogram in
the past two years? \| female \| 2018 \|
+-------------+---------------------------------------------------------------------------------------------------------------------------------------------------------
-------------------------------------+-------------+------------------+

+-------------+---------------------------------------------------------------------------------------------------------------------------------------------------------
-------------------------------------+-------------+------------------+
\| \_RFPAP34  \| Women respondents aged 50-74 who have had a pap test in
the past two years? \| female \| 2018 \|
+-------------+---------------------------------------------------------------------------------------------------------------------------------------------------------
-------------------------------------+-------------+------------------+
