---
title: "Measurement of sex and gender identity in BRFSS"
author: "Jenine K Harris, Odalis Hernandez"
date: "9/15/2021"
output: 
  word_document:
      reference_docx: "template.docx"
bibliography: references.bib
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE,
                      warning = FALSE)
source("data-clean-and-analysis.R")
```

### Background

In 2021, a Gallup poll found that .6% or 60 out of every 1000 Americans,
reported that they identify as transgender [@lgbtgallup]. The percentage
varied by age group, with the cohort born in 1997-2002 reporting 1.8% of
survey participants identified as transgender compared to 1.2% in the
previous cohort and just .3% in the cohort born before 1946. Transgender
people are individuals whose biological sex, or sex recorded at birth,
differs from their gender identity [@galupo2016labels; @howerton2021].
Gender identity is how a person identifies and conceptualizes their own
gender and is different from biological sex and sexual orientation
[@galupo2016labels; @tate2020undoing]. Sex, sexual orientation, and
gender identity all have implications for health, health care, and
health outcomes [@legato2016consideration; @agenor2016sexual;
@gonzales2016comparison; @mcclure2022gender].

Conducting research to understand factors that influence health, health
care, and health outcomes is important for creating guidelines,
policies, and programs that protect and improve health. Given the roles
of sex, sexual orientation, and gender identity in health, many social
science and health surveys seek to measure sex, and a growing number
also collect sexual orientation and gender identity information. Until
recently, measurement of gender identity was unusual and measurement of
sex was often done by an interviewer recording the sex they thought the
participant was based on voice or other cues [@bauer2017transgender] or
asking "what is your sex?" and recording either male or female. In
addition, researchers have frequently treated gender as a synonym for
sex [@krieger2003genders; @johnson2009better; @tate2013two].

For decades there have been calls and attempts to improve the
measurement of sex, gender, and sexual orientation [@krieger2003genders;
@tate2013two]. In the meantime, researchers are utilizing data sets that
include less accurate ways of measuring these concepts. For example, the
Behavioral Risk Factor Surveillance Survey (BRFSS) collects data on
health behaviors, health conditions, and the use of health-related
services from U.S residents. The Centers for Disease Control and
Prevention collect BRFSS participant data each month over telephone. It
is considered the world's largest health survey system, collecting data
from over 400,000 individuals yearly [@brfss].

Until 2015, BRFSS did not ask participants about gender identity. In
2015 the BRFSS added an optional question about gender identity to the
survey and states could decide whether or not to include the gender
identity question when conducting data collection. The gender identity
question that was included in the BRFSS from 2015 until 2020 (the most
recent year of data released) was: "Do you consider yourself to be
transgender?" The follow-up question for those who answered yes was, "Do
you consider yourself to be 1. Male-to-female, 2. Female-to-male, or 3.
Gender nonconforming?"

Although a gender identity question was added, the wording of the sex
question did not make any distinction between sex and gender for several
years. In 2015, the interviewer selected the sex of the participant
given the prompt: "Indicate sex of respondent. Ask only if necessary."
In 2016, the prompt changed to, "Are you..." with the options of Male,
Female, and Refused and a note, "Note: This may be populated from
information derived from screening, household enumeration. However,
interviewer should not make judgement on sex of respondent." So, 2016
was the first year of the BRFSS that the interviewer was not permitted
to guess the sex of the participant.

In 2017 the question remained, "Are you..." with responses of Male,
Female, and Refused. The note for interviewers changed to all capital
letters, presumably for emphasis, "INTERVIEWER NOTE: THIS QUESTION MUST
BE ASKED EVEN IF INTERVIEWER HAD PREVIOUSLY ENTERED SEX IN THE SCREENING
QUESTIONS. IT WILL NOT BE ASKED OF PERSONS WHO HAVE SELF-IDENTIFIED SEX
IN LL HOUSEHOLD ENUMERATION." There was also a second note, "CATI NOTE:
THIS QUESTION MAY BE POPULATED BY LANDLINE HOUSEHOLD ENUMERATION ONLY.
IT MAY NOT BE POPULATED BY INTERVIEWER ASSIGNMENT OF SEX DURING THE
SCREENING FOR CELL PHONE OR PERSONS LIVING IN COLLEGE HOUSING."

In 2018, the interviewer asked, "What is your sex? or What was your sex
at birth? Was it..." and gave the possible responses of Male and Female.
It is unclear from the BRFSS documentation how interviewers selected one
of the two questions. In 2019, the interviewer asked, "Are you male or
female?". In 2019 there was also a separate "Sex at Birth Optional
Module" that included the question, "What was your sex at birth? Was it
male or female?" The separate sex at birth question was asked of a
subset of about 65 thousand participants and a combined sex variable was
created that classified people based on their response to the sex at
birth question if asked and, if sex at birth was not asked or if the
reply was "Don't know/Not Sure" or "Refused," used the response to the
sex question. In the 2020 survey administration, the 2019 sex at birth
prompt was adopted as the sex question for all participants, "What was
your sex at birth? Was it male or female?"

In a recent paper [@howerton2021], Howerton and Harris found that BRFSS
participants from 2015 to 2019 who reported having the same transgender
gender identity sometimes had different answers for the sex question.
For example, some participants who identified as female-to-male
transgender identity reported their sex as male and others reported
their sex as female. Likewise, among those who identified as
male-to-female transgender identity, some reported their sex to be male
and others reported their sex to be female. This has likely led to
numerous published scientific papers examining sex differences or using
sex as a covariate and potentially including participants whose sex
recorded at birth was something other than their sex as recorded by the
BRFSS.

The implications of how sex is measured impacts which questions
participants are asked and how multiple variables are computed in the
BRFSS. For example, the BRFSS variable for heavy drinking codes a male
as a heavy drinker at 14 or more alcoholic drinks per week and a female
as a heavy drinker at 7 or more alcoholic drinks per week. Questions
related to pregnancy and pregnancy or birth-related health conditions
are only asked of those who report being female and questions about
human papilloma virus (HPV) screening and vaccination and birth control
type and behavior differ for those who select male or female sex.
Likewise, questions about mammograms, breast exams, and prostate
specific antigen tests are only asked of people who report being in a
specific sex category.

We sought to better understand the relationship between sex and gender
identity responses in the BRFSS data from 2015 (the first year of the
gender identity question) to the most recently released year of data in
2020. Specifically we answer three questions: (1) What is the
relationship between sex and gender identity responses for participants
over this six year period?; (2) How has the relationship between
measured sex and gender identity changed as the sex question changed
over this six year period?; and (3) Did the versions of the question
that specified "at birth" result in different patterns of sex and gender
identity reporting compared to other years?

### Methods

We imported the BRFSS data sets from 2015 through 2020 directly from the
Centers for Disease Control and Prevention website. We used R
statistical software version 4.1.1 [@rstats], the RStudio (version
1.4.1717) interactive development environment [@rstudio], and a function
from version 2.4.3 of the haven package [@haven] to import the data. We
used the tidyverse [@wickham2017] software package to manage and
visualize the data and the table1 package to compute descriptive
statistics [@table1].

### Results

##### What is the relationship between sex and gender identity responses for participants over this six year period?

The majority of those who reported being male-to-female transgender also
reported their sex as male each year (Figure 1). Likewise, the majority
of those who reported being female-to-male transgender reported their
sex as female. With the exception of 2017, there was a higher percentage
of male-to-female transgender participants reporting themselves to be
male sex compared to the percentage of female-to-male transgender
participants reporting themselves to be female.

*Figure 1. Distribution of reported sex among BRFSS participants who
reported being transgender (2015-2020).*

```{r}
sixBars
```

##### How has the relationship between measured sex and gender identity changed as the sex question changed over this six year period?

Although the sex question asked of participants changed multiple times
from 2015 to 2020, the percentage in each transgender group choosing
male and female did not change notably for either group (Figure 2).

```{r}
overTime
```

##### Did the versions of the question that specified "at birth" result in different patterns of sex and gender identity reporting compared to other years?

For the transgender male-to-female group, the 2018 and 2020 had a
slightly different pattern of responses compared to the years just
before these (more in line with the 2015 responses), with more
participants selecting male sex in 2018 than in 2017 and in 2020
compared to 2019. For the transgender female-to-male group, 2018
question also marked a shift in responses with more participants
choosing male sex than previous years. This small change was maintained
in 2019 but there was a slight increase in female sex responses in 2020.
The "at birth" wording appears to have had more of an impact on
participants identifying as male-to-female transgender than on those
identifying as female-to-male transgender. It is notable that, for both
of the transgender groups (female-to-male and male-to-female) there was
an increase in selection of male in response to the sex question in
2018, the first year of the "at birth" wording.

The structure of the 2019 survey provides some additional information
about differences in response patterns when including or not including
"at birth" in a sex question. In 2019, there was more variability in
responses among participants who were asked their sex (Figure 3A)
compared to participants who were asked their sex at birth (Figure 3B).
However, when the sex at birth version of the question was adopted for
everyone in 2020 (Figure 3C), the distribution of responses was
somewhere between the responses from those who responded to the sex
question in 2019 and those who responded to the sex at birth question in
2019.

```{r}
cowplot::plot_grid(plot_grid(sexVar + theme(legend.position = "none"),
                   sexAtBirthVar + theme(legend.position = "none"),
                   sexAtBirthVar2020 + theme(legend.position = "none"), nrow = 1,
                   labels = c("A", "B", "C")),
                   legend, ncol = 1, rel_heights = c(1,.1))
```

### Discussion

We found that about 75% of participants in the 2015 to 2020 BRFSS who
identify as transgender male-to-female report male sex while about 25%
report female sex. During the same time frame, participants who identify
as transgender female-to-male report female sex a little over 60% of the
time and report male sex just under 40% of the time. These values were
relatively consistent over the six administrations of the survey despite
multiple changes in how sex was measured. The one exception to this was
in the group of transgender male-to-female participants who answered the
optional module on gender identity in 2019. When asked their sex at
birth, about 85% of this group chose male. The value of the sex variable
is used in computing some health related variables (e.g., heavy
drinking) and in routing participants to questions about birth control,
HPV vaccines, cancer screenings, and other preventive health behaviors.

The lack of *accurate* population-level epidemiological data for
transgender people is problematic [@dente2019]. For example,
population-level research is lacking on breast cancer among transgender
people, even though many take hormone replacement therapy which could
influence breast cancer risk [@dente2019]. In another example
[@tabaac2018], three years of BRFSS data were used to determine
colonoscopy and mammography screenings among transgender people but
neglected to mention the possible mismatch between the participant
response to the sex variable and the types of screenings recommended.

In addition, the BRFSS calculated variables contain a measure of heavy
drinking and a measure of binge drinking. Both of these measures
classify participants based on sex and do not consider gender identity
in classification; the threshold for heavy drinking is quite different
for males (14 drinks per week) compared to females (7 drinks per week),
while the binge drinking definition differs by a single drink per
episode (5 for males, 4 for females). Given participant responses for
sex vary within a single gender identity, these thresholds may not be
accurately capturing heavy or binge drinking. For example, although
Azagba et al. [@azagba2019] reported the sex distribution for each
transgender group in the first table of their paper, they did not
include sex as a covariate in logistic regression models computing the
odds of binge drinking for different transgender groups compared to
cisgender groups. Perhaps not surprising, the findings suggest that
female-to-male transgender participants have a lower odds of binge
drinking compared to cisgender males and male-to-female transgender
participants had higher odds of binge drinking compared to cisgender
females [@azagba2019]. Given the importance of understanding alcohol
consumption in the transgender population [@gilbert2018], a way of
classifying hazardous consumption that takes into consideration physical
attributes rather than relying solely on sex could improve accuracy of
this measure and thus improve the quality of research on alcohol use
among transgender people.

There are at least two possible remedies for the BRFSS and other large
scale surveys to reduce the current limitations on accurate
population-level data on transgender health. First, the BRFSS (and other
large scale surveys) could ask all of the questions currently asked of
only male sex or only female sex for any person who identifies as
transgender. In this way we can ensure more population-level information
about important cancer screenings, health conditions, and health
behaviors among transgender people. Second, researchers including the
sex variable in their work should pay attention to measures of gender
identity if they are available and, if gender identity measures are not
available, might report on this limitation given that reported sex is
likely a mixture of sex recorded at birth and gender identity.

Although our work examined the BRFSS, ineffective data collected by
other types of instruments, like medical records, are also impacting
transgender health. For example, although most transgender men still
have a cervix, they are less likely to be screened for cervical cancer
compared to their cisgender counterparts [@gatos2018] with one the
reasons being that some health systems base screening invitations on the
sex listed health records [@connolly2020].

Measurement of sex and gender identity is evolving [@beischel2022].

### References
