---
title: "Measurement of sex and gender identity in BRFSS"
author: "Jenine K Harris, Odalis Hernandez"
date: "9/15/2021"
output: 
  word_document:
      reference_docx: "template.docx"
bibliography: references.bib
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE,
                      warning = FALSE)
source("data-clean-and-analysis.R")
```

### Background

In 2021, a Gallup poll found that .6% or 60 out of every 1000 Americans,
reported that they identify as transgender [@lgbtgallup]. The percentage
varied by age group, with the cohort born in 1997-2002 reporting 1.8% of
survey participants identified as transgender compared to 1.2% in the
previous cohort and just .3% in the cohort born before 1946. Transgender
people are individuals whose biological sex, or sex recorded at birth,
differs from their gender identity [@galupo2016labels; @howerton2021].
Gender identity is how a person identifies and conceptualizes their own
gender and is different from biological sex and sexual orientation
[@galupo2016labels; @tate2020undoing]. Sex, sexual orientation, and
gender identity all have implications for health, health care, and
health outcomes [@legato2016consideration; @agenor2016sexual;
@gonzales2016comparison; @mcclure2022gender].

Conducting research to understand factors that influence health, health
care, and health outcomes is important for creating guidelines,
policies, and programs that protect and improve health. Given the roles
of sex, sexual orientation, and gender identity in health, many social
science and health surveys seek to measure sex, and a growing number
also collect sexual orientation and gender identity information. Until
recently, measurement of gender identity was unusual and measurement of
sex was often done by an interviewer recording the sex they thought the
participant was based on voice or other cues [@bauer2017transgender] or
asking "what is your sex?" and recording either male or female. In
addition, researchers have frequently treated gender as a synonym for
sex [@krieger2003genders; @johnson2009better; @tate2013two].

For decades there have been calls and attempts to improve the
measurement of sex, gender, and sexual orientation [@krieger2003genders;
@tate2013two]. In the meantime, researchers are utilizing data sets that
include less accurate ways of measuring these concepts. For example, the
Behavioral Risk Factor Surveillance Survey (BRFSS) collects data on
health behaviors, health conditions, and the use of health-related
services from U.S residents. The Centers for Disease Control and
Prevention collect BRFSS participant data each month over telephone. It
is considered the world's largest health survey system, collecting data
from over 400,000 individuals yearly [@brfss].

Until 2014, BRFSS did not ask participants about gender identity. In
2015 the BRFSS added an optional question about gender identity to the
survey and states could decide whether or not to include the gender
identity question when conducting data collection. The gender identity
question that was included in the BRFSS from 2014 until 2020 (the most
recent year of data released) was: "Do you consider yourself to be
transgender?" The follow-up question for those who answered yes was, "Do
you consider yourself to be 1. Male-to-female, 2. Female-to-male, or 3.
Gender nonconforming?"

Although a gender identity question was added, the wording of the sex
question did not make any distinction between sex and gender for several
years. In 2014 and 2015, the interviewer selected the sex of the
participant given the prompt: "Indicate sex of respondent. Ask only if
necessary." In 2016, the prompt changed to, "Are you..." with the
options of Male, Female, and Refused and a note, "Note: This may be
populated from information derived from screening, household
enumeration. However, interviewer should not make judgement on sex of
respondent." So, 2016 was the first year of the BRFSS that the
interviewer was not permitted to guess the sex of the participant.

In 2017 the question remained, "Are you..." with responses of Male,
Female, and Refused. The note for interviewers changed to all capital
letters, presumably for emphasis, "INTERVIEWER NOTE: THIS QUESTION MUST
BE ASKED EVEN IF INTERVIEWER HAD PREVIOUSLY ENTERED SEX IN THE SCREENING
QUESTIONS. IT WILL NOT BE ASKED OF PERSONS WHO HAVE SELF-IDENTIFIED SEX
IN LL HOUSEHOLD ENUMERATION." There was also a second note, "CATI NOTE:
THIS QUESTION MAY BE POPULATED BY LANDLINE HOUSEHOLD ENUMERATION ONLY.
IT MAY NOT BE POPULATED BY INTERVIEWER ASSIGNMENT OF SEX DURING THE
SCREENING FOR CELL PHONE OR PERSONS LIVING IN COLLEGE HOUSING."

In 2018, the interviewer asked, "What is your sex? or What was your sex
at birth? Was it..." and gave the possible responses of Male and Female.
It is unclear from the BRFSS documentation how interviewers selected one
of the two questions. In 2019, the interviewer asked, "Are you male or
female?". In 2019 there was also a separate "Sex at Birth Optional
Module" that included the question, "What was your sex at birth? Was it
male or female?" The separate sex at birth question was asked of a
subset of about 65 thousand participants and a combined sex variable was
created that classified people based on their response to the sex at
birth question if asked and, if sex at birth was not asked or if the
reply was "Don't know/Not Sure" or "Refused," used the response to the
sex question. In the 2020 survey administration, the 2019 sex at birth
prompt was adopted as the sex question for all participants, "What was
your sex at birth? Was it male or female?"

In a recent paper [@howerton2021], Howerton and Harris found that BRFSS
participants from 2015 to 2019 who reported having the same transgender
gender identity sometimes had different answers for the sex question.
For example, some participants who identified as female-to-male
transgender identity reported their sex as male and others reported
their sex as female. Likewise, among those who identified as
male-to-female transgender identity, some reported their sex to be male
and others reported their sex to be female. This has likely led to
numerous published scientific papers examining sex differences or using
sex as a covariate and potentially including participants whose sex
recorded at birth was something other than their sex as recorded by the
BRFSS.

The implications of how sex is measured impacts which questions
participants are asked and how multiple variables are computed in the
BRFSS. For example, the BRFSS variable for heavy drinking codes a male
as a heavy drinker at 14 or more alcoholic drinks per week and a female
as a heavy drinker at 7 or more alcoholic drinks per week. Questions
related to pregnancy and pregnancy or birth-related health conditions
are only asked of those who report being female and questions about
human papilloma virus (HPV) screening and vaccination and birth control
type and behavior differ for those who select male or female sex.
Likewise, questions about mammograms, breast exams, and prostate
specific antigen tests are only asked of people who report being in a
specific sex category.

We sought to better understand the relationship between sex and gender
identity responses in the BRFSS data from 2014 (the first year of the
gender identity question) to the most recently released year of data in
2020. Specifically we answer three questions: (1) What is the
relationship between sex and gender identity responses for participants
over this six year period?; (2) How has the relationship between
measured sex and gender identity changed as the sex question changed
over this six year period?; and (3) Did the versions of the question
that specified "at birth" result in different patterns of sex and gender
identity reporting compared to other years?

### Methods

We imported the BRFSS data sets from 2014 through 2020 directly from the
Centers for Disease Control and Prevention website. We used R
statistical software version 4.1.1 [@rstats], the RStudio (version
1.4.1717) interactive development environment [@rstudio], and a function
from version 2.4.3 of the haven package [@haven] to import the data. We
used the tidyverse [@wickham2017] software package to manage and
visualize the data and the table1 package to compute descriptive
statistics [@table1].

### Results

##### What is the relationship between sex and gender identity responses for participants over this six year period?

The majority of those who reported being male-to-female transgender also
reported their sex as male each year (Figure 1). Likewise, the majority
of those who reported being female-to-male transgender reported their
sex as female. With the exception of 2017, there was a higher percentage
of male-to-female transgender participants reporting themselves to be
male sex compared to the percentage of female-to-male transgender
participants reporting themselves to be female.

*Figure 1. Distribution of reported sex among BRFSS participants who
reported being transgender (2014-2020).*

```{r}
reposition_legend(sixBars +  
                                     facet_wrap(.~yearLong,nrow = 2,
                                                labeller = label_wrap_gen(20)), 
                                   position = "bottom",
                                   panel='panel-4-2')
```

##### How has the relationship between measured sex and gender identity changed as the sex question changed over this six year period?

Although the sex question asked of participants changed multiple times
from 2014 to 2020, the percentage in each transgender group choosing
male and female remained relatively stable for both groups (Figure 2).

```{r}
overTime
```

##### Did the versions of the question that specified "at birth" result in different patterns of sex and gender identity reporting compared to other years?

The transgender male-to-female group ranged from
`r round(100*min(percTransTib$percentTranYr[percTransTib$TRNSGNDR == "Trans male-to-female" & percTransTib$SEX == 'male']),0)`%
to
`r round(100*max(percTransTib$percentTranYr[percTransTib$TRNSGNDR == "Trans male-to-female" & percTransTib$SEX == 'male']),0)`%
male sex with no apparent pattern given the question wording changes.
The transgender female-to-male group ranged from
`r round(100*min(percTransTib$percentTranYr[percTransTib$TRNSGNDR == "Trans female-to-male" & percTransTib$SEX == 'female']),0)`%
to
`r round(100*max(percTransTib$percentTranYr[percTransTib$TRNSGNDR == "Trans female-to-male" & percTransTib$SEX == 'female']),0)`%
female sex with a pattern of decline in responding "female" to the sex
question from 2014 to 2020.

The structure of the 2019 survey provides some additional information
about differences in response patterns when including or not including
"at birth" in a sex question. In 2019, there was more variability in
responses among participants who were asked their sex (Figure 3A)
compared to participants who were asked their sex at birth (Figure 3B).
However, when the sex at birth version of the question was adopted for
everyone in 2020 (Figure 3C), the distribution of responses was
somewhere between the responses from those who responded to the sex
question in 2019 and those who responded to the sex at birth question in
2019.

```{r}
cowplot::plot_grid(plot_grid(sexVar + theme(legend.position = "none"),
                   sexAtBirthVar + theme(legend.position = "none"),
                   sexAtBirthVar2020 + theme(legend.position = "none"), nrow = 1,
                   labels = c("A", "B", "C")),
                   legend, ncol = 1, rel_heights = c(1,.1))
```

### Discussion

We found that about `r round(100*percTrans[1,1], 0)`% of participants in
the 2014 to 2020 BRFSS who identify as transgender male-to-female report
male sex while about `r 100-round(100*percTrans[1,1], 0)`% report female
sex. During the same time frame, participants who identify as
transgender female-to-male report female sex
`r round(100*percTrans[2,2], 0)`% of the time and report male sex
`r 100-   round(100*percTrans[2,2], 0)`% of the time. These values were
relatively consistent over the seven administrations of the survey;
there were some minor shifts year-to-year in the sex distribution in the
male-to-female group and a relatively steady increase in reporting male
sex in the female-to-male group. One exception to the relatively
consistent percentages in the male-to-female transgender group was for
the participants who answered the optional module on gender identity in
2019. When asked their sex at birth,
`r round(100*percTrans2019[1,1], 0)`% of this group chose male. The
value of the sex variable is used in computing some health related
variables (e.g., heavy drinking) and in routing participants to
questions about birth control, HPV vaccines, cancer screenings, and
other preventive health behaviors.

The lack of *accurate* population-level epidemiological data for
transgender people is problematic [@dente2019]. For example,
population-level research is lacking on breast cancer among transgender
people, even though many take hormone replacement therapy which could
influence breast cancer risk [@dente2019]. In one example [@tabaac2018],
three years of BRFSS data were used to determine colonoscopy and
mammography screenings among transgender people but neglected to discuss
the possible mismatch between the participant response to the sex
variable and the types of screenings recommended. Likewise, another
study using BRFSS data relied on the *Are you transgender?* question and
the sex question to classify people as transgender men and transgender
women [@tabaac2018] before analysis of the associations between a five
category variable (cisgender men, cisgender women, transgender men,
transgender women, and gender non-conforming) and several cancer
screening tests including pap and PSA tests. While both transgender men
and transgender women were asked about mammogram, transgender women were
not asked about pap tests and transgender men were not asked bout PSA.
In this case, with the way these groups were classified by the authors,
those who reported male sex but transgender status female-to-male (n =
212) would have been asked the PSA screening question and not the pap
screening question. Likewise, those who reported female sex but
male-to-female transgender status (n = 319) would have been asked the
pap screening question but not the PSA. So, out of the 1,779 transgender
male-to-female or female-to-male participants, this strategy likely
missed 212 people who likely still had a cervix [@gatos2018] but were
not asked about a pap test and 319 of people who potentially had a
prostate but were not asked about a PSA.

Similar inadequate data storage in medical records is exacerbating the
problem. For example, although most transgender men still have a cervix,
they are less likely to be screened for cervical cancer compared to
their cisgender counterparts [@gatos2018] with one of the reasons being
that some health systems base screening invitations on the sex listed
health records [@connolly2020]. Lack of screening for transgender people
can lead to later diagnoses and more severe disease [@Bertoncelli2022,
@Yang2018]. It is clear that our existing tools are inadequate for
collecting, storing, and using data effectively for transgender people.

In addition ineffective direct measurement of sex, the BRFSS includes
computed variables that rely on the value of the sex variable to
classify participants. Two of these variables are are *heavy drinking*
and *binge drinking*. These measures classify participants as heavy
drinkers or binge drinkers based on the number of drinks they report
having in a week and their sex. They do not consider gender identity in
classification; the threshold for heavy drinking is quite different for
males (14 drinks per week) compared to females (7 drinks per week),
while the binge drinking definition differs by a single drink per
episode (5 for males, 4 for females). Given participant responses for
sex vary within a single gender identity, these thresholds may not be
accurately capturing heavy or binge drinking. In one example of how this
impacts research, Azagba et al. [@azagba2019] did not include sex as a
covariate in logistic regression models computing the odds of binge
drinking for different transgender groups compared to cisgender groups.
Perhaps not surprising, the findings suggested that female-to-male
transgender participants have a lower odds of binge drinking compared to
cisgender males and male-to-female transgender participants had higher
odds of binge drinking compared to cisgender females [@azagba2019].
Given the importance of understanding alcohol consumption in the
transgender population [@gilbert2018], a way of classifying hazardous
consumption that takes into consideration physical attributes rather
than relying solely on sex could improve accuracy of this measure and
thus improve the quality of research on alcohol use among transgender
people.

There are at least two possible ways to reduce the current limitations
on accurate population level data for transgender people. First, the
BRFSS (and other data collection and storage tools) could build in the
flexibility to ask all of the questions currently asked of only male sex
or only female sex for any person who identifies as transgender. In this
way we can ensure more accurate information about important cancer
screenings, health conditions, and health behaviors among transgender
people. Second, researchers including the sex variable in their work
should pay attention to measures of gender identity if they are
available and, if gender identity measures are not available, might
report on this limitation given that data on reported sex for
transgender people is likely a mixture of sex recorded at birth and
gender identity.

More flexible and accurate ways of measurement of sex and gender
identity are also necessary to protect and improve the health of
transgender people [@beischel2022; @Bauer2017]. Measures including
transgender identity currently include single questions and two-step
questions typically asking about sex at birth and gender identity
separately, like the BRFSS 2020 administration [@measurin2022]. While
the two-step questions provide more detailed information and has been
suggested by some groups as a best practice [@conron2014identifying],
they are not perfect. Numerous studies suggest drawbacks including a
small number of people who would fit the definition of transgender but
who do not identify as transgender and a small number of cisgender
participants responding that they are transgender because they do not
understand the terminology [@measurin2022]. One study also found
evidence that some transgender people may misunderstand trangender
terminology, with a small number of transgender women self-identifying
as transgender men [@Costa2022].

While there does not seem to be consensus yet on the best way to measure
sex and gender, scientists from different fields continue to work on it.
For example, a 2017 study [@Bauer2017] examined different ways of
measuring sex and gender and consulted with experts to identify a set of
three questions, "What sex were you assigned at birth, meaning on your
original birth certificate?"; "Which best describes your current gender
identity?"; and, for those whose gender identity differs from birth sex,
"What gender do you currently live as in your day-to-day life?" In
contrast, Thornton et al. [@thornton2022] argue that, as a step toward
best practice for gender and sex data collection, ethical measurement
could skip the sex question as a filter and use questions that are
relevant to a study topic like "Do you have a prostate?" or "Are you
breastfeeding?" Whether one of these options is the key remains to be
seen. What is clear, though, is that ineffective data collection,
storage, and use is a threat to the health of transgender people. More
accurate, flexible, and inclusive measurement along with data systems
that are more flexible for data collection, storage, and use are
strategies that could improve the health and well-being of transgender
people.

### References
