---
title: "Measurement of sex and gender identity in BRFSS"
author: "Jenine K Harris, Odalis Hernandez"
date: "9/15/2021"
output: word_document
bibliography: references.bib
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = FALSE, message = FALSE,
                      warning = FALSE)
```

```{r}
# remove the eval = FALSE to run
# TO CREATE ANALYTIC DATA SET, UNCOMMENT AND RUN THIS CODE CHUNK
# bring in files from online BRFSS 2015 to 2018 data
library(haven)
# #
# #
# # 2015 data
  temp <- tempfile(fileext = ".zip")
  download.file(url  = "https://www.cdc.gov/brfss/annual_data/2015/files/LLCP2015XPT.zip", destfile = temp)
  BRFSS_2015 <- read_xpt(file = temp)
# # #
# # 2016 data
 temp <- tempfile(fileext = ".zip")
 download.file(url  = "https://www.cdc.gov/brfss/annual_data/2016/files/LLCP2016XPT.zip", destfile = temp)
 BRFSS_2016 <- read_xpt(file = temp)
# #
# # # 2017 data
 temp <- tempfile(fileext = ".zip")
 download.file(url  = "https://www.cdc.gov/brfss/annual_data/2017/files/LLCP2017XPT.zip", destfile = temp)
 BRFSS_2017 <- read_xpt(file = temp)
# #
# 2018 data
temp <- tempfile(fileext = ".zip")
download.file(url  = "https://www.cdc.gov/brfss/annual_data/2018/files/LLCP2018XPT.zip", destfile = temp)
BRFSS_2018 <- read_xpt(file = temp)
#
# # 2019 data
temp <- tempfile(fileext = ".zip")
download.file(url  = "https://www.cdc.gov/brfss/annual_data/2019/files/LLCP2019XPT.zip", destfile = temp)
BRFSS_2019 <- read_xpt(file = temp)
#
# 2020 data
temp <- tempfile(fileext = ".zip")
download.file(url  = "https://www.cdc.gov/brfss/annual_data/2020/files/LLCP2020XPT.zip", destfile = temp)
BRFSS_2020 <- read_xpt(file = temp)
```

```{r}
# remove the eval = FALSE to run
# # select only the variables to use in the analysis (e.g. changes 300+ variables to only needed for analysis)

#
library(tidyverse)
 BRFSS_2015 <- BRFSS_2015 %>%
   select(SEX, TRNSGNDR, `_RFBING5`, `_RFDRHV5`, DIABETE3,
          PREGNANT, HADMAM, HOWLONG, HADPAP2, LASTPAP2, HPVTEST,
          HPLSTTST, HADHYST2, PROFEXAM, LENGEXAM, PCPSAAD2, PCPSADI1,
          PCPSARE1, PSATEST1, PSATIME, PCPSARS1, PCPSADE1, PCDMDECN,
          `_DRNKWEK`) %>%
    mutate(year = "2015")
#
 BRFSS_2016 <- BRFSS_2016 %>%
   select(SEX, `_DRNKWEK`, TRNSGNDR, `_RFDRHV5`, PREGNANT, HADMAM,
          HOWLONG, HADPAP2, LASTPAP2, HPVTEST, HPLSTTST, HADHYST2,
          PROFEXAM, LENGEXAM, PCPSAAD2, PCPSADI1, PCPSARE1, PSATEST1,
          PSATIME, PCPSARS1, PCPSADE1, PCDMDECN, `_RFMAM2Y`, `_MAM5021`,
          `_RFPAP33`, `_RFPSA21`)%>%
    mutate(year = "2016")
#
 BRFSS_2017 <- BRFSS_2017 %>%
   select(SEX, `_DRNKWEK`, TRNSGNDR, `_RFDRHV5`, PREGNANT, PFPPRVN2,
          TYPCNTR7, NOBCUSE6)%>%
    mutate(year = "2017")
#
 BRFSS_2018 <- BRFSS_2018 %>%
  select(SEX1, `_DRNKWEK`,  TRNSGNDR, `_LLCPWT`)%>%
    mutate(year = "2018")
# #
 BRFSS_2019 <- BRFSS_2019 %>%
   select(`_MICHD`, `_SEX`, EDUCA, `_BMI5CAT`, `_DRNKWK1`, EXERANY2, `_RACE`,
          `_AGE80`, `_SMOKER3`, TRNSGNDR, `_LLCPWT`)%>%
    mutate(year = "2019")
# 
 BRFSS_2020 <- BRFSS_2020 %>%
    mutate(year = "2020") %>%
    rename(SEX = `_SEX`)
#
# # renamed SEX1 in 2018 to SEX like all the other years
 BRFSS_2018 <- BRFSS_2018 %>%
   rename(SEX = SEX1)
 BRFSS_2019 <- BRFSS_2019 %>%
   rename(SEX = `_SEX`)
#
 # renamed AVEDRNK3 to AVEDRNK2 in 2019 like all the other years
BRFSS_2019 <- BRFSS_2019 %>%
  rename(`_DRNKWEK` = `_DRNKWK1`)

# 
# 
# 
# # # combine all years into 1 dataset
BRFSS_all <- rbind(select(BRFSS_2015, SEX, TRNSGNDR, year),
                   select(BRFSS_2016, SEX, TRNSGNDR, year),
                   select(BRFSS_2017, SEX, TRNSGNDR, year),
                   select(BRFSS_2018, SEX, TRNSGNDR, year),
                   select(BRFSS_2019, SEX, TRNSGNDR, year),
                   select(BRFSS_2020, SEX, TRNSGNDR, year))
# 
# # save BRFSS_all dataset as a .csv file so every time i want to work with, I just read that file
# # in instead of running all steps above
# write.csv(BRFSS_all, file = "brfss_all.csv")
```

```{r}
#BRFSS_all <- read.csv("brfss_all.csv")
# add labels
BRFSS_all_clean <- BRFSS_all %>% 
   mutate(SEX = recode_factor(SEX, 
                              '1' = 'male',
                              '2' = 'female')) %>% 
   mutate(TRNSGNDR = recode_factor(TRNSGNDR,
                                   '1' = 'Trans male-to-female',
                                   '2' = 'Trans female-to-male',
                                   '3' = 'Trans gender nonconforming',
                                   '4' = 'No',
                                   '7' = NA_character_,
                                   '9' = NA_character_)) %>% 
   filter(TRNSGNDR %in% c("Trans male-to-female",
                          "Trans female-to-male")) %>% 
  mutate(yearLong = recode_factor(year,
                              '2015' = '2015: Indicate sex of respondent. Ask only if necessary.',
                              '2016' = '2016: Are you... (Male, Female, Refused)',
                              '2017' = '2017: Are you male or female?',
                              '2018' = '2018: What is your sex? or What was your sex at birth? Was it...',
                              '2019' = '2019: Are you male or female? or What was your sex at birth? Was it male or female?',
'2020' = '2020: What was your sex at birth? Was it male or female?')) %>% 
  drop_na() %>% 
   droplevels()
```

### Background

In 2021, a Gallup poll found that .6% or 60 out of every 1000 Americans,
reported that they identify as transgender [@lgbtgallup]. The percentage
varied by age group, with the cohort born in 1997-2002 reporting 1.8% of
survey participants identified as transgender compared to 1.2% in the
previous cohort and just .3% in the cohort born before 1946. Transgender
people are individuals whose biological sex, or sex recorded at birth,
differs from their gender identity [@galupo2016labels; @howerton2021].
Gender identity is how a person identifies and conceptualizes their own
gender and is different from biological sex and sexual orientation
[@galupo2016labels; @tate2020undoing]. Sex, sexual orientation, and
gender identity all have implications for health, health care, and
health outcomes [@legato2016consideration; @agenor2016sexual;
@gonzales2016comparison; @mcclure2022gender].

Conducting research to understand factors that influence health, health
care, and health outcomes is important for the development of policy and programs that protect and improve health. Given the roles of sex, sexual orientation, and gender
identity in health, many social science and health surveys seek to
measure sex, and a growing number also collect sexual orientation and
gender identity information. Until recently, measurement of gender
identity was unusual and measurement of sex was often done by an
interviewer recording the sex they thought the participant was based on
voice or other physical cues [@bauer2017transgender] or asking "what is
your sex?" and recording either male or female. In addition, researchers
have frequently treated gender as a synonym for sex
[@krieger2003genders; @johnson2009better; @tate2013two].

For decades there have been calls and attempts to improve the measurement of sex,
gender, and sexual orientation [@krieger2003genders; @tate2013two]. In
the meantime, researchers are utilizing data sets that include less
accurate ways of measuring these concepts. For example, the Behavioral
Risk Factor Surveillance Survey (BRFSS) collects data on health behaviors,
health conditions, and the use of health-related services from U.S
residents. The Centers for Disease Control and Prevention collect BRFSS participant data  each month over telephone. It is considered the world's largest
health survey system, collecting data from over 400,000 individuals
yearly [@brfss].

Until 2015, BRFSS did not ask participants about gender identity. In
2015 the BRFSS added an optional question about gender identity to the
survey and states could decide whether or not to include the gender
identity question when conducting data collection. The gender identity
question that was included in the BRFSS from 2015 until 2020 (the most
recent year of data released) was: "Do you consider yourself to be
transgender?" The follow-up question for those who answered yes was, "Do
you consider yourself to be 1. Male-to-female, 2. Female-to-male, or 3.
Gender nonconforming?"

Although a gender identity question was added, the wording of the sex
question did not make any distinction between sex recorded and gender
for several years. In 2015, the interviewer selected the sex of
the participant given the prompt: "Indicate sex of respondent. Ask only if necessary." In 2016, the prompt changed to, "Are you..." with the options of Male, Female, and Refused and a note, "Note: This may be populated from information derived from screening, household enumeration. However, interviewer should not make judgement on sex of respondent." So, 2016 was the first year of the BRFSS that the interviewer was not permitted to guess the sex of the participant.

In 2017 the question remained, "Are you..." with responses of Male, Female, and Refused. The note for interviewers changed to all capital letters, presumably for emphasis, "INTERVIEWER NOTE: THIS QUESTION MUST BE ASKED EVEN IF INTERVIEWER HAD PREVIOUSLY ENTERED SEX IN THE SCREENING QUESTIONS. IT WILL NOT BE ASKED OF PERSONS WHO HAVE SELF-IDENTIFIED SEX IN LL HOUSEHOLD ENUMERATION." There was also a second note, "CATI NOTE: THIS QUESTION MAY BE POPULATED BY LANDLINE HOUSEHOLD ENUMERATION ONLY. IT MAY NOT BE POPULATED BY INTERVIEWER ASSIGNMENT OF SEX DURING THE SCREENING FOR CELL PHONE OR PERSONS LIVING IN COLLEGE HOUSING."

In 2018,
the interviewer asked, "What is your sex? or What was your sex at birth?
Was it..." and gave the possible responses of Male and Female. It is unclear from the BRFSS documentation how interviewers selected one of the two questions. In 2019,
the interviewer asked, "Are you male or female?". In 2019 there was also a
separate "Sex at Birth Optional Module" that included the question,
"What was your sex at birth? Was it male or female?" The separate sex at birth question was asked of a subset of about 65 thousand participants and a combined sex variable was created that classified people based on their response to the sex at birth question if asked and, if sex at birth was not asked or if the reply was "Don't know/Not Sure" or "Refused," used the response to the sex question. In the 2020 survey
administration, the 2019 sex at birth prompt was adopted as the
sex question for all participants, "What was your sex at birth? Was it male or female?"

In a recent paper [@howerton2021], Howerton and Harris found that BRFSS
participants from 2015 to 2019 who reported having the same transgender
gender identity sometimes had different answers for the sex question. For example,
some participants who identified as female-to-male transgender identity
reported their sex as male and others reported their sex as female.
Likewise, among those who identified as male-to-female transgender
identity, some reported their sex to be male and others reported their
sex to be female. This has likely led to numerous published scientific
papers examining sex differences or using sex as a covariate and potentially including participants
whose sex recorded at birth was something other than their sex as
recorded in the BRFSS.

The implications of how sex is measured also impacts multiple computed
variables in the BRFSS that rely on the sex variable as part of the
calculation. For example, the BRFSS variable for heavy drinking codes a
male as a heavy drinker at 14 or more alcoholic drinks per week and a
female as a heavy drinker at 7 or more alcoholic drinks per week.
Questions related to pregnancy and pregnancy or birth-related health
conditions are only asked of those who report being female and questions
about human
papilloma virus (HPV) screening and vaccination and birth control type and behavior differ for those who select male
or female sex. Likewise, questions about mammograms, breast exams, and prostate specific
antigen tests are only asked of people who report being in a specific
sex category.

We sought to better understand the relationship between sex and gender
identity responses in the BRFSS data from 2015 (the first year of the gender
identity question) to the most recently released year of data in 2020.
Specifically we answer three questions: (1) What is the relationship
between sex and gender identity responses for participants over this six
year period?; (2) How has the relationship between measured sex and
gender identity changed as the sex question changed over this six year
period?; and (3) Did the versions of the question that specified "at
birth" result in different patterns of sex and gender identity reporting
compared to other years?

### Methods

We imported the BRFSS data sets from 2015 through 2020 directly from the
Centers for Disease Control and Prevention website. We used R
statistical software version 4.1.1 [@rstats], the RStudio (version
1.4.1717) interactive development environment [@rstudio], and a function
from version 2.4.3 of the haven package [@haven] to import the data. We
used the tidyverse [@wickham2017] software package to manage and
visualize the data and the table1 package to compute descriptive
statistics [@table1].

### Results

##### What is the relationship between sex and gender identity responses for participants over this six year period?

The majority of those who reported being male-to-female transgender also reported their sex as male each year (Figure 1). Likewise, the majority of those who reported being female-to-male transgender reported their sex as female. With the exception of 2017, there was a higher percentage of male-to-female transgender participants reporting themselves to be male sex compared to the percentage of female-to-male transgender participants reporting themselves to be female.

*Figure 1. Distribution of reported sex among BRFSS participants who reported being transgender (2015-2020).*

```{r}
BRFSS_all_clean %>% 
  drop_na() %>% 
  group_by(TRNSGNDR, SEX, yearLong) %>% 
  count() %>% 
  group_by(TRNSGNDR, yearLong) %>% 
  mutate(percYear = 100 * n / sum(n)) %>% 
  ggplot(aes(x = TRNSGNDR, fill = SEX, y = percYear)) +
  geom_col(position = "dodge") +
  #coord_flip() +
  scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 10)) +
  facet_wrap(~yearLong, 
             labeller = label_wrap_gen(20)) +
  scale_fill_manual(values = c("gray60", "black")) +
  theme_minimal() +
  labs(x = "Gender identity", y = "Percent in each sex category",
       fill = "Sex category")  

```

##### How has the relationship between measured sex and gender identity changed as the sex question changed over this six year period?

Although the sex question asked of participants changed multiple times from 2015 to 2020, the percentage in each transgender group choosing male and female did not change notably for either group (Figure 2). 

```{r}
BRFSS_all_clean %>% 
  drop_na() %>% 
  mutate(year = as.numeric(as.character(year))) %>% 
  group_by(TRNSGNDR, SEX, year) %>% 
  count() %>% 
  group_by(TRNSGNDR, year) %>% 
  mutate(percYear = 100 * n / sum(n)) %>% 
  ggplot(aes(color = SEX, x = year, y = percYear)) +
  geom_path() +
  geom_point() +
  #scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 10)) +
  facet_wrap(~TRNSGNDR, nrow = 2) +
  scale_color_manual(values = c("gray60", "black")) +
  theme_minimal() +
  labs(x = "Year", y = "Percent in each sex category",
       color = "Sex category")  + 
  scale_x_continuous(labels=c("2015","2016","2017", "2018", "2019","2020")) +
  ylim(c(0,100))

```

##### Did the versions of the question that specified "at birth" result in different patterns of sex and gender identity reporting compared to other years?

For the transgender male-to-female group, the 2018 and 2020 had a slightly different pattern of responses compared to the years just before these (more in line with the 2015 responses), with more participants selecting male sex in 2018 than in 2017 and in 2020 compared to 2019. For the transgender female-to-male group, 2018 question also marked a shift in responses with more participants choosing male sex than previous years. This small change was maintained in 2019 but there was a slight increase in female sex responses in 2020. The "at birth" wording appears to have had more of an impact on participants identifying as male-to-female transgender than on those identifying as female-to-male transgender. It is notable that, for both of the transgender groups (female-to-male and male-to-female) there was an increase in selection of male in response to the sex question in 2018. 

The structure of the 2019 survey provides some additional information about differences in response patterns when including or not including "at birth" in a sex question. In 2019, there was more variability in responses among participants who were asked their sex (A) compared to participants who were asked their sex at birth (B). However, when the sex at birth version of the question was adopted in 2020 (Figure 3C), the distribution of responses was somewhere between the responses from those who responded to the sex question in 2019 ans those who responded to the sex at birth question in 2019. 

```{r}
temp <- tempfile(fileext = ".zip")
download.file(url  = "https://www.cdc.gov/brfss/annual_data/2019/files/LLCP2019XPT.zip", destfile = temp)
BRFSS_2019_raw <- read_xpt(file = temp)

BRFSS_2019_clean <- BRFSS_2019_raw %>% 
  mutate(SEXVAR = recode_factor(SEXVAR,
                                '1' = 'male',
                                '2' = 'female')) %>% 
  mutate(BIRTHSEX = recode_factor(BIRTHSEX,
                                '1' = 'male',
                                '2' = 'female',
                                '7' = NA_character_,
                                '9' = NA_character_)) %>% 
  mutate(TRNSGNDR = recode_factor(TRNSGNDR,
                                   '1' = 'Trans male-to-female',
                                   '2' = 'Trans female-to-male',
                                   '3' = 'Trans gender nonconforming',
                                   '4' = "No",
                                   '7' = NA_character_,
                                   '9' = NA_character_))

table(BRFSS_2019_clean$SEXVAR, BRFSS_2019_clean$BIRTHSEX)
table(BRFSS_2019_clean$TRNSGNDR, BRFSS_2019_clean$BIRTHSEX)
```
```{r}
library(cowplot)

sexVar <- BRFSS_2019_clean %>% 
  drop_na(TRNSGNDR, SEXVAR) %>% 
  filter(TRNSGNDR != "No" & TRNSGNDR != "Trans gender nonconforming") %>% 
  group_by(TRNSGNDR, SEXVAR) %>% 
  count() %>% 
  group_by(TRNSGNDR) %>% 
  mutate(percSex = 100 * n / sum(n)) %>% 
  ggplot(aes(x = TRNSGNDR, fill = SEXVAR, y = percSex)) +
  geom_col(position = "dodge") +
  #coord_flip() +
  scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 10)) +
  # facet_wrap(~yearLong, 
  #            labeller = label_wrap_gen(20)) +
  scale_fill_manual(values = c("gray60", "black")) +
  theme_minimal() +
  labs(x = "Gender identity", y = "Percent in each\ntransgender category (sex 2019)",
       fill = "Sex category")  +
  ylim(c(0, 100))

sexAtBirthVar <- BRFSS_2019_clean %>% 
  drop_na(TRNSGNDR, BIRTHSEX) %>% 
  filter(TRNSGNDR != "No" & TRNSGNDR != "Trans gender nonconforming") %>% 
  group_by(TRNSGNDR, BIRTHSEX) %>% 
  count() %>% 
  group_by(TRNSGNDR) %>% 
  mutate(percSex = 100 * n / sum(n)) %>% 
  ggplot(aes(x = TRNSGNDR, fill = BIRTHSEX, y = percSex)) +
  geom_col(position = "dodge") +
  #coord_flip() +
  scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 10)) +
  # facet_wrap(~yearLong, 
  #            labeller = label_wrap_gen(20)) +
  scale_fill_manual(values = c("gray60", "black")) +
  theme_minimal() +
  labs(x = "Gender identity", y = "Percent in each\ntransgender category (sex at birth 2019)",
       fill = "Sex category") +
  ylim(c(0, 100))

sexAtBirthVar2020 <- BRFSS_all_clean %>% 
  filter(year == 2020) %>% 
  drop_na(TRNSGNDR, SEX) %>% 
  filter(TRNSGNDR != "No" & TRNSGNDR != "Trans gender nonconforming") %>% 
  group_by(TRNSGNDR, SEX) %>% 
  count() %>% 
  group_by(TRNSGNDR) %>% 
  mutate(percSex = 100 * n / sum(n)) %>% 
  ggplot(aes(x = TRNSGNDR, fill = SEX, y = percSex)) +
  geom_col(position = "dodge") +
  #coord_flip() +
  scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 10)) +
  # facet_wrap(~yearLong, 
  #            labeller = label_wrap_gen(20)) +
  scale_fill_manual(values = c("gray60", "black")) +
  theme_minimal() +
  labs(x = "Gender identity", y = "Percent in each \ntransgender category (sex at birth 2020)",
       fill = "Sex category") +
  ylim(c(0, 100))

legend <- get_legend(
  # create some space to the left of the legend
  sexVar +
     guides(color = guide_legend(nrow = 1)) +
    theme(legend.position = "bottom"))


cowplot::plot_grid(plot_grid(sexVar + theme(legend.position = "none"),
                   sexAtBirthVar + theme(legend.position = "none"),
                   sexAtBirthVar2020 + theme(legend.position = "none"), nrow = 1,
                   labels = c("A", "B", "C")),
                   legend, ncol = 1, rel_heights = c(1,.1))
```


### Discussion

We found that about 75% of participants in the 2015 to 2020 BRFSS who identify as transgender male-to-female report male sex while about 25% report female sex. For participants during the same time frame who identify as transgender female-to-male report female sex a little over 60% of the time and report male sex just under 40% of the time. These values were relatively consistent over the six administrations of the survey despite multiple changes in how sex was measured. The one exception to this was in the group of transgender male-to-female participants who answered the optional module in 2019. When asked their sex at birth, about 85% of this group chose male. The value of the sex variable is used in computing some health related variables (e.g., heavy drinking) and in routing participants to questions about birth control, HPV vaccines, cancer screenings, and other preventive health behaviors. In order to better understand and protect the health of transgender people, accurate measurement of biological sex could be useful.

[paragraph about evidence-based measurement of bio sex]

[paragraph about calculated variables and routing, can BRFSS route trans-male-to-female to breast cancer screen]

[paragraph suggesting researchers pay attention]





```{r}
# basic table exploring data
library(table1)
table1(~ year | TRNSGNDR * SEX, data = BRFSS_all_clean)
```



```{r}
library(descr)

chisqtrans <- function(yr) {
  CrossTable(BRFSS_all_clean$SEX[BRFSS_all_clean$year == yr],
           BRFSS_all_clean$TRNSGNDR[BRFSS_all_clean$year == yr],
           chisq = TRUE,
           sresid = TRUE,
          prop.c = FALSE,
          prop.t = FALSE,
          prop.chisq = FALSE,
          expected = TRUE)
}

chisqtrans(2015)
chisqtrans(2016)
chisqtrans(2017)
chisqtrans(2018)
chisqtrans(2019)
chisqtrans(2020)

```

+-------------+-------------------------+-------------+-------------+
| Q u e s t i | Question                | A s k e d o | Y ea r( s)  |
| o n i d     |                         | f . . .     |             |
+=============+=========================+=============+=============+
| \           |                         | c o m p u t | 20 15       |
| \           |                         | e d         |             |
| \_ D R N K  |                         |             |             |
| W K 1       |                         |             |             |
+-------------+-------------------------+-------------+-------------+
| \           | Binge drinkers (males   | c o m p u t | 20 15 , 20  |
| \           | having five or more     | e d         | 18          |
| \_ R F B I  | drinks on one occasion, |             |             |
| N G 5       | females having four or  |             |             |
|             | more drinks on one      |             |             |
|             | occasion)               |             |             |
+-------------+-------------------------+-------------+-------------+
| \           | Heavy drinkers (adult   | c o m p u t | 2 01 5, 2   |
| \           | men having more than 14 | e d         | 01 6, 20 17 |
| \_ R F D R  | drinks per week and     |             |             |
| H V 5       | adult women having more |             |             |
|             | than 7 drinks per week) |             |             |
+-------------+-------------------------+-------------+-------------+
| \           | Heavy drinkers (adult   | c o m p u t | 20 18       |
| \           | men having more than 14 | e d         |             |
| \_ R F D R  | drinks per week and     |             |             |
| H V 6       | adult women having more |             |             |
|             | than 7 drinks per week) |             |             |
+-------------+-------------------------+-------------+-------------+

+-------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
+-------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
---------------+-------------+------------------+ \| DIABETE3 \| (Ever
told) you have diabetes (If "Yes" and respondent is female, ask "Was
this only when you were pregnant?". If Respondent says pre-diabetes or
borderline diabetes, use response code 4.) \| female \| 2015 2018 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+

\| PREGNANT \| To your knowledge, are you now pregnant? \| female \|
2015, 2016, 2017, 2018, 2019 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| HADMAM \| A mammogram is an x-ray of each breast to look for breast
cancer. Have you ever had a mammogram? \| female \| 2015, 2016, 2018,
2019 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| HOWLONG \| How long has it been since you had your last mammogram? \|
female \| 2015, 2016, 2018, 2019 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| HADPAP2 \| A Pap test is a test for cancer of the cervix. Have you
ever had a Pap test? \| female \| 2015, 2016, 2018, 2019 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| LASTPAP2 \| How long has it been since you had your last Pap test? \|
female \| 2015, 2016, 2018, 2019 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| HPVTEST \| An HPV test is sometimes given with the Pap test for
cervical cancer screening. Have you ever had an HPV test? \| female \|
2015, 2016, 2018, 2019 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| HPLSTTST \| How long has it been since you had your last HPV test? \|
female \| 2015, 2016, 2018, 2019 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| HADHYST2 \| Have you had a hysterectomy? (A hysterectomy is an
operation to remove the uterus (womb).) \| female \| 2015, 2016, 2018,
2019 \|

+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| PROFEXAM \| A clinical breast exam is when a doctor, nurse, or other
health professional feels the breast for lumps. Have you ever had a
clinical breast exam? \| female \| 2015, 2016, 2018 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| LENGEXAM \| How long has it been since your last breast exam? \|
female \| 2015, 2016, 2018 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| PCPSAAD2 \| Has a doctor, nurse, or other health professional EVER
talked with you about the advantages of the PSA test? \| male \| 2015,
2016 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+

+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| PCPSARE1 \| Has a doctor, nurse, or other health professional EVER
recommended that you have a PSA test? \| male \| 2015, 2016, 2018, 2019
\|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+

+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| PCDMDECN \| Who made the decision with you? (Mark all that apply) \|
male \| 2015, 2016 \|
+-------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

+-------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
---------------+-------------+------------------+ \| \_RFMAM2Y  \| Women
respondents aged 40+ who have had a mammogram in the past two years \|
computed \| 2016 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| \_MAM5021  \| Women respondents aged 50-74 who have had a mammogram
in the past two years \| computed \| 2016 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| \_RFPAP33  \| Women respondents aged 21-65 who have had a pap test in
the past three years \| computed \| 2016 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| \_RFPSA21  \| Male respondents aged 40+ who have had a PSA test in
the past 2 years \| computed \| 2016 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| PFPPRVN2  \| Did you or your partner do anything the last time you
had sex to keep you from getting pregnant? \| female \| 2017, 2019 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| TYPCNTR7  \| What did you or your partner do the last time you had
sex to keep you from getting pregnant? \| female \| 2017 \|
+-------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
---------------+-------------+------------------+ \| TYPCNTR8  \| What
did you or your partner do the last time you had sex to keep you from
getting pregnant? \| female \| 2019 \|
+-------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
---------------+-------------+------------------+ \| NOBCUSE6  \| What
was your main reason for not doing anything the last time you had sex to
keep you from getting pregnant? \| female \| 2017 \|
+-------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
---------------+-------------+------------------+ \| NOBCUSE7  \| What
was your main reason for not doing anything the last time you had sex to
keep you from getting pregnant? \| female \| 2019 \|
+-------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
---------------+-------------+------------------+
---------------------+-------------+------------------+ \| CNCRTYP1  \|
With your most recent diagnoses of cancer, what type of cancer was it?
\| female, male \| 2018, 2019 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| CSRVDOC1  \| What type of doctor provides the majority of your health
care? \| female \| 2018, 2019 \|
+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+

+-------------+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------+------------------+
\| \_RFMAM21  \| Women respondents aged 40+ who have had a mammogram in
the past two years? \| female \| 2018 \|
+-------------+---------------------------------------------------------------------------------------------------------------------------------------------------------
-------------------------------------+-------------+------------------+

+-------------+---------------------------------------------------------------------------------------------------------------------------------------------------------
-------------------------------------+-------------+------------------+
\| \_RFPAP34  \| Women respondents aged 50-74 who have had a pap test in
the past two years? \| female \| 2018 \|
+-------------+---------------------------------------------------------------------------------------------------------------------------------------------------------
-------------------------------------+-------------+------------------+
