---
title: "Measurement of sex and gender identity in BRFSS"
author: "Jenine K Harris, Odalis Hernandez"
date: "3/21/2022"
output: 
  word_document:
      reference_docx: "template.docx"
bibliography: references.bib
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE,
                      warning = FALSE)
source("data-clean-and-analysis.R")
```

# Abstract

# Background

In 2021, a Gallup poll found that .6% or 60 out of every 1000 Americans,
reported that they identify as transgender [@lgbtgallup\]. The
percentage varied by age group, with the cohort born in 1997-2002
reporting 1.8% of survey participants identified as transgender compared
to 1.2% in the previous cohort and just .3% in the cohort born before
1946. Transgender people are individuals whose biological sex, or sex
recorded at birth, differs from their gender identity
\[@galupo2016labels; @howerton2021\]. Gender identity is how a person
identifies and conceptualizes their own gender and is different from
biological sex and sexual orientation \[@galupo2016labels;
@tate2020undoing\]. Sex, sexual orientation, and gender identity all
have implications for health, health care, and health outcomes
\[@legato2016consideration; @agenor2016sexual; @gonzales2016comparison;
@mcclure2022gender].

Conducting research to understand factors that influence health, health
care, and health outcomes is important for creating guidelines,
policies, and programs that protect and improve health. Given the roles
of sex, sexual orientation, and gender identity in health, many social
science and health surveys seek to measure sex, and a growing number
also collect sexual orientation and gender identity information. Until
recently, measurement of gender identity was not commonly included as a
survey item and measurement of sex was often done by an interviewer
guessing the sex of the participant [@bauer2017transgender] or asking
"what is your sex?" and recording either male or female. Researchers
have frequently treated gender as a synonym for sex
[@krieger2003genders; @johnson2009better; @tate2013two].


For decades there have been calls and attempts to improve the
measurement of sex, gender identity, and sexual orientation
[@krieger2003genders; @tate2013two]. In the meantime, researchers are
utilizing data sets that include many different ways of measuring these
concepts. One of these data sets that is commonly used is the Behavioral
Risk Factor Surveillance Survey (BRFSS), which collects data on health
behaviors, health conditions, and the use of health-related services
from U.S residents. The Centers for Disease Control and Prevention
collect BRFSS participant data each month over telephone. It is
considered the world's largest health survey system, collecting data
from over 400,000 individuals yearly [@brfss].

Until 2014, BRFSS did not include questions about gender identity. In
2014, the sexual orientation and gender identity (SOGI) module was
offered and 19 states included it in the survey this first year. The
number of states including the SOGI module increased over time to 33
states in the most recent data release in 2020 [@brfss2020sogi]. The
gender identity question included in the SOGI is: "Do you consider
yourself to be transgender?" The follow-up question for those who
answered yes was, "Do you consider yourself to be 1. Male-to-female, 2.
Female-to-male, or 3. Gender nonconforming?" Prior to offering the
optional SOGI module to all states, a few states included their own
gender identity questions in some years (e.g., Massachusetts in 2007 to
2009 [@Conron2012]).

Although a gender identity question was added, the wording of the sex
question did not make any distinction between sex and gender for several
years. In 2014 and 2015, the interviewer selected the sex of the
participant given the prompt: "Indicate sex of respondent. Ask only if
necessary." In 2016, the prompt changed to, "Are you..." with the
options of Male, Female, and Refused and a note, "Note: This may be
populated from information derived from screening, household
enumeration. However, interviewer should not make judgement on sex of
respondent." So, 2016 was the first year of the BRFSS that the
interviewer was not permitted to guess the sex of the participant.

In 2017 the question remained, "Are you..." with responses of Male,
Female, and Refused. The note for interviewers changed to all capital
letters, presumably for emphasis, "INTERVIEWER NOTE: THIS QUESTION MUST
BE ASKED EVEN IF INTERVIEWER HAD PREVIOUSLY ENTERED SEX IN THE SCREENING
QUESTIONS. IT WILL NOT BE ASKED OF PERSONS WHO HAVE SELF-IDENTIFIED SEX
IN LL HOUSEHOLD ENUMERATION." There was also a second note, "CATI NOTE:
THIS QUESTION MAY BE POPULATED BY LANDLINE HOUSEHOLD ENUMERATION ONLY.
IT MAY NOT BE POPULATED BY INTERVIEWER ASSIGNMENT OF SEX DURING THE
SCREENING FOR CELL PHONE OR PERSONS LIVING IN COLLEGE HOUSING."

In 2018, the BRFSS interviewers asked, "What is your sex? or What was
your sex at birth? Was it..." and gave the possible responses of Male
and Female. It is unclear from the BRFSS documentation how interviewers
selected one of the two questions. In 2019, the interviewer asked, "Are
you male or female?". In 2019 there was also a separate "Sex at Birth
Optional Module" that included the question, "What was your sex at
birth? Was it male or female?" The separate sex at birth question was
asked of a subset of about 65 thousand participants and a combined sex
variable was created that classified people based on their response to
the sex at birth question if asked and, if sex at birth was not asked or
if the reply was "Don't know/Not Sure" or "Refused," used the response
to the sex question. In the 2020 survey administration, the 2019 sex at
birth prompt was adopted as the sex question for all participants, "What
was your sex at birth? Was it male or female?"

In a recent paper [@howerton2021], Howerton and Harris found that BRFSS
participants from 2015 to 2019 who reported having the same transgender
gender identity sometimes had different answers for the sex question.
For example, some participants who identified as female-to-male
transgender identity reported their sex as male and others reported
their sex as female. Likewise, among those who identified as
male-to-female transgender identity, some reported their sex to be male
and others reported their sex to be female. This has likely led to
numerous published scientific papers examining sex differences or using
sex as a covariate and potentially including participants whose sex
recorded at birth was something other than their self-reported sex
recorded in the BRFSS data.

The implications of how sex is measured impacts which questions
participants are asked and how multiple variables are computed in the
BRFSS. For example, the BRFSS variable for heavy drinking codes a male
as a heavy drinker at 14 or more alcoholic drinks per week and a female
as a heavy drinker at 7 or more alcoholic drinks per week. Questions
related to pregnancy and pregnancy or birth-related health conditions
are only asked of those who report being female and questions about
human papilloma virus (HPV) screening and vaccination and birth control
type and behavior differ for those who select male or female sex.
Likewise, questions about mammograms, breast exams, and prostate
specific antigen tests are only asked of people who report being in a
specific sex category.

We sought to better understand the relationship between sex and gender
identity responses in the BRFSS data from 2014 (the first year of the
gender identity question) to the most recently released year of data in
2020. Specifically we answer three questions: (1) What is the
relationship between sex and gender identity responses for participants
over this six year period?; (2) How has the relationship between
measured sex and gender identity changed as the sex question changed
over this six year period?; and (3) Did the versions of the question
that specified "at birth" result in different patterns of sex and gender
identity reporting compared to other years?

# Methods

We imported the BRFSS data sets from 2014 through 2020 directly from the
Centers for Disease Control and Prevention website. We used R
statistical software version 4.1.1 [@rstats], the RStudio (version
1.4.1717) interactive development environment [@rstudio], and a function
from version 2.4.3 of the haven package [@haven] to import the data. We
used the tidyverse [@wickham2017] software package to manage and
visualize the data and the table1 package to compute descriptive
statistics [@table1].

# Results

## What is the relationship between sex and gender identity responses for participants over this six year period?

The majority of those who reported being male-to-female transgender also
reported their sex as male each year (Figure 1). Likewise, the majority
of those who reported being female-to-male transgender reported their
sex as female. With the exception of 2017, there was a higher percentage
of male-to-female transgender participants reporting themselves to be
male sex compared to the percentage of female-to-male transgender
participants reporting themselves to be female.

*Figure 1. Distribution of reported sex among BRFSS participants who
reported being transgender (2014-2020).*

```{r}
reposition_legend(sixBars +  
                                     facet_wrap(.~yearLong,nrow = 2,
                                                labeller = label_wrap_gen(20)), 
                                   position = "bottom",
                                   panel='panel-4-2')
```

## How has the relationship between measured sex and gender identity changed as the sex question changed over this six year period?

Although the sex question asked of participants changed multiple times
from 2014 to 2020, the percentage in each transgender group choosing
male and female remained relatively stable for both groups (Figure 2).

```{r}
overTime
```

## Did the versions of the question that specified "at birth" result in different patterns of sex and gender identity reporting compared to other years?

The transgender male-to-female group ranged from
`r round(100*min(percTransTib$percentTranYr[percTransTib$TRNSGNDR == "Trans male-to-female" & percTransTib$SEX == 'male']),0)`%
to
`r round(100*max(percTransTib$percentTranYr[percTransTib$TRNSGNDR == "Trans male-to-female" & percTransTib$SEX == 'male']),0)`%
male sex with no apparent pattern given the question wording changes.
The transgender female-to-male group ranged from
`r round(100*min(percTransTib$percentTranYr[percTransTib$TRNSGNDR == "Trans female-to-male" & percTransTib$SEX == 'female']),0)`%
to
`r round(100*max(percTransTib$percentTranYr[percTransTib$TRNSGNDR == "Trans female-to-male" & percTransTib$SEX == 'female']),0)`%
female sex with a pattern of decline in responding "female" to the sex
question from 2014 to 2020.

The structure of the 2019 survey provides some additional information
about differences in response patterns when including or not including
"at birth" in a sex question. In 2019, there was more variability in
responses among participants who were asked their sex (Figure 3A)
compared to participants who were asked their sex at birth (Figure 3B).
However, when the sex at birth version of the question was adopted for
everyone in 2020 (Figure 3C), the distribution of responses was
somewhere between the responses from those who responded to the sex
question in 2019 and those who responded to the sex at birth question in
2019.


```{r}
cowplot::plot_grid(plot_grid(sexVar + theme(legend.position = "none"),
                   sexAtBirthVar + theme(legend.position = "none"),
                   sexAtBirthVar2020 + theme(legend.position = "none"), nrow = 1,
                   labels = c("A", "B", "C")),
                   legend, ncol = 1, rel_heights = c(1,.1))
```

# Discussion

We found that about `r round(100*percTrans[1,1], 0)`% of participants in
the 2014 to 2020 BRFSS who identify as transgender male-to-female report
male sex while about `r 100-round(100*percTrans[1,1], 0)`% report female
sex. During the same time frame, participants who identify as
transgender female-to-male report female sex
`r round(100*percTrans[2,2], 0)`% of the time and report male sex
`r 100-   round(100*percTrans[2,2], 0)`% of the time. These values were
relatively consistent over the seven administrations of the survey;
there were some minor shifts year-to-year in the sex distribution in the
male-to-female group and a relatively steady increase in reporting male
sex in the female-to-male group. One exception was for the participants
who answered the optional module on gender identity in 2019. When asked
their sex at birth, `r round(100*percTrans2019[1,1], 0)`% of the
male-to-female transgender group chose male. The value of the sex
variable is used in computing some health related variables (e.g., heavy
drinking) and in routing participants to questions about birth control,
HPV vaccines, cancer screenings, and other preventive health behaviors.

The lack of *accurate* population-level epidemiological data for
transgender people is problematic [@dente2019]. For example,
population-level research is lacking on breast cancer among transgender
people, even though many take hormone replacement therapy which could
influence breast cancer risk [@dente2019]. In another example
[@tabaac2018], three years of BRFSS data were used to determine
colonoscopy and mammography screenings among transgender people but
neglected to discuss the possible mismatch between the participant
response to the sex variable and the types of screenings recommended.
Likewise, a study using BRFSS data relied on the *Are you transgender?*
question and the sex question to classify people as transgender men and
transgender women [@tabaac2018] before analysis of the associations
between a five category variable (cisgender men, cisgender women,
transgender men, transgender women, and gender non-conforming) and
several cancer screening tests including pap and PSA tests. While both
transgender men and transgender women were asked about mammogram,
transgender women were not asked about pap tests and transgender men
were not asked bout PSA. In this case, with the way these groups were
classified by the authors, those who reported male sex but transgender
status female-to-male, and therefore may have a cervix and not a
prostate (n = 212), would have been asked the PSA screening question and
not the pap screening question. Likewise, those who reported female sex
but male-to-female transgender status, and therefore may have a prostate
and not a cervix (n = 319), would have been asked the pap screening
question but not the PSA. So, out of the 1,779 transgender
male-to-female or female-to-male participants, this strategy likely
missed 212 people who likely had a cervix [@gatos2018] but were not
asked about a pap test and 319 of people who potentially had a prostate
but were not asked about a PSA test.

Similar inadequate data storage in medical records is exacerbating the
problem. For example, although most transgender men still have a cervix,
they are less likely to be screened for cervical cancer compared to
their cisgender counterparts [@gatos2018] with one of the reasons being
that some health systems base screening invitations on the sex listed in
medical records [@connolly2020]. Lack of screening for transgender
people can lead to later diagnoses and more severe disease
[@Bertoncelli2022] [@Yang2018]. It is clear that our existing tools are
inadequate for collecting, storing, and using data effectively for
transgender people.

In addition ineffective direct measurement of sex, the BRFSS includes
computed variables that rely on the value of the sex variable to
classify participants. Two of these variables are are *heavy drinking*
and *binge drinking*. These measures classify participants as heavy
drinkers or binge drinkers based on the number of drinks they report
having in a week and their sex. They do not consider gender identity in
classification; the threshold for heavy drinking is quite different for
males (14 drinks per week) compared to females (7 drinks per week),
while the binge drinking definition differs by a single drink per
episode (5 for males, 4 for females). Given participant responses for
sex vary within a single gender identity, these thresholds may not be
accurately capturing heavy or binge drinking for transgender
participants. In one example of how this impacts research, Azagba et al.
[@azagba2019] did not include sex as a covariate in logistic regression
models computing the odds of binge drinking for different transgender
groups compared to cisgender groups. Perhaps not surprising, the
findings suggested that female-to-male transgender participants have a
lower odds of binge drinking compared to cisgender males and
male-to-female transgender participants had higher odds of binge
drinking compared to cisgender females [@azagba2019]. Given the
importance of understanding alcohol consumption in the transgender
population [@gilbert2018], a way of classifying hazardous consumption
that takes into consideration physical attributes rather than relying
solely on self-reported sex could improve accuracy of this measure and
thus improve the quality of research on alcohol use among transgender
people.

There are at least two possible ways to reduce the current limitations
on accurate population level data for transgender people. First, the
BRFSS (and other data collection and storage tools) could build in the
flexibility to ask all of the questions currently asked of only male sex
or only female sex for any person who identifies as transgender. In this
way we can ensure more accurate information about important cancer
screenings, health conditions, and health behaviors among transgender
people. Second, researchers including the sex variable in their work
should pay attention to measures of gender identity if they are
available and, if gender identity measures are not available, might
report on this limitation given that data on reported sex for
transgender people is likely a mixture of sex recorded at birth and
gender identity.

More flexible and accurate ways of measurement of sex and gender
identity are also necessary to protect and improve the health of
transgender people [@beischel2022; @Bauer2017]. Current ways of asking
gender identity include single questions with a transgender option and
two-step questions typically asking about sex at birth and gender
identity separately [@measurin2022]. While the two-step questions
provide more detailed information and have been suggested by some groups
as a best practice [@conron2014identifying], they are not perfect.
Numerous studies suggest drawbacks including a small number of people
who would fit the definition of transgender but who do not identify as
transgender and a small number of cisgender participants responding that
they are transgender because they do not understand the terminology
[@measurin2022]. One study also found evidence that some transgender
people may misunderstand trangender terminology, with a small number of
transgender women self-identifying as transgender men [@Costa2022]. In
our study, the two-step question using the "at birth" terminology in
2020 did not result in notable changes in the reporting of sex for those
who were transgender from prior years.

While there does not seem to be consensus yet on the best way to measure
sex and gender identity as distinct, scientists from different fields
continue to examine the data, test options, and develop guidelines. For
example, a 2017 study [@Bauer2017] examined different ways of measuring
sex and gender and consulted with experts to identify a set of three
questions, "What sex were you assigned at birth, meaning on your
original birth certificate?"; "Which best describes your current gender
identity?"; and, for those whose gender identity differs from birth sex,
"What gender do you currently live as in your day-to-day life?" A 2021
paper suggests gender identity be measured by two questions, "Which of
the following best describes your gender identity? Woman, Man,
Non-binary, genderfluid, or gender non-conforming, or Other (please
specify)" and "Do you identify as transgender? Yes, No, Other (please
specify)" [@kosciesza2022]. Likewise, the National Academies of
Sciences, Engineering, and Medicine recommended a two question set for
sex and gender identity, "What sex were you assigned at birth, on your
original birth certificate? Female, Male, Don't know, Prefer not to
answer" and "What is your current gender? Female, Male, Transgender, (if
respondent is American Indian / Alaskan Native) Two-Spirit, I use a
different term (free test), Don't know, Prefer not to answer"
[@measurin2022]. In contrast, Thornton et al. [@thornton2022] argue
that, as a step toward best practice for gender and sex data collection,
ethical measurement could skip the sex question as a filter and use
questions that are relevant to a study topic like "Do you have a
prostate?" or "Are you breastfeeding?"

Whether one of these options is the key to more effective and accurate
data collection remains to be seen. What is clear, though, is that
ineffective data collection, storage, and use is a threat to the health
of transgender people and researchers using these data should be aware
of its limitations. More accurate, flexible, and inclusive measurement
along with data systems that are more flexible for data collection,
storage, and use are strategies that could improve the health and
well-being of transgender people.

# References
